---
title: "Real data results: Marker genes for cell types"
subtitle: "Cell type comparison: PT_S1 v. C_TAL"
author: "Florica"
date: "2025-09-01"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE,
                      message = FALSE)
```

Supplementary table 5 of "An atlas of healthy and injured cell states and niches in the human kidney" by Lake et. al., contains a list of marker genes for major kidney cell types. 
We map these kidney cell types to the kidney cell types present in our data. 
In particular, we find marker genes for the cell types "C_TAL" (cortical thick ascending limb of Henle) and "PT_S1" (proximal tubule segment 1). 
These two cell types are well represented in our data as they have a high number of cells across all samples. 
With our subsequent analyses we aim not to discover any new biology findings, but rather to perform a "sanity check" that our algorithm performs as expected. 
In particular, we test to see which genes are differentially expressed between "C_TAL" cells and "PT_S1" cells within each of the three kidney conditions (Control, HKD, and DKD). 
The marker genes should have large test statistics and thus be highly statistically significant in this analysis. 
Additionally, as we are studying marker genes, standard non-spatial GLM models should also find similar results and rank the same marker genes as highly statistically significant.

Concretely, we focus on the top 3000 genes, ranked by variance. 
We run both the TESSERA algorithm, with the Poisson-Leroux model, and non-spatial generalized linear models (Poisson and Negative Binomial GLMs) on the same set of genes.
For each gene, we fit a model with fixed effects for the interaction of cell-type and disease, no main effects, and no intercept, and fixed effects for sample that are nested within disease; note that this model is exactly equivalent to the model with traditional main effects for cell-type and disease, an intercept, and interactions, and that the results are identical across both models. 

Concretely, we test 
$H_0: \beta_{C\_TAL, \textrm{condition}} == \beta_{PT\_S1, \textrm{condition}}$,
$H_1: \beta_{C\_TAL, \textrm{condition}} \neq \beta_{PT\_S1, \textrm{condition}}$,
for each condition.
We construct a Wald test statistic and use our threshold/fitting procedure to obtain p-values. 
In this case, the contrasts of interest are simply differences between the two entries of $\beta$. 

In this setting, in an interaction-only model, the contrasts we test are
$\beta_{C\_TAL, \textrm{condition}} - \beta_{PT\_S1, \textrm{condition}} == 0$ 
for each condition. 
In a model with main effects, an intercept, and a reference level, we would test
$(\beta_{\textrm{intercept}} + \beta_{\textrm{condition}} + \beta_{C\_TAL} + \beta_{C\_TAL, \textrm{condition}}) - (\beta_{\textrm{intercept}} + \beta_{\textrm{condition}} + \beta_{PT\_S1} + \beta_{PT\_S1, \textrm{condition}}) = (\beta_{C\_TAL} + \beta_{C\_TAL, \textrm{condition}}) - (\beta_{PT\_S1} + \beta_{PT\_S1, \textrm{condition}}) == 0$.

Both of these models are equivalent, in that we may uniquely isolate any combination of cell-type and condition with different linear combinations of the entries of $\beta$.
If the condition or "C_TAL" were reference levels above, we would simply drop those terms in the contrast equation above, e.g., if Control were the condition and Control and "C_TAL" were reference levels, we would write
$(\beta_{PT\_S1})== 0$
for the contrast;
however, if the condition were DKD instead, we would write
$(\beta_{C\_TAL, \textrm{DKD}}) - (\beta_{PT\_S1} + \beta_{PT\_S1, \textrm{DKD}}) == 0$.

Moreover, fitting both models on the data, for both the GLM models and the Poisson-Leroux models, leads to identical residuals, identical contrast values, and identical Wald statistics, as expected. 
The advantage of the interaction-only model is the simplicity of forming contrasts for the questions of interest here, as the lack of reference levels simplifies interpretation. 

```{r}
# Library
library(openxlsx)
library(dplyr)
library(forcats)
library(ggplot2)
library(patchwork)
library(TESSERA)

# Paths
processed_data_path <- "../../data/processed/"
raw_data_path <- "../../data/raw/"
figure_path <- "../../figures/real/DE_marker_genes/"

# Model type
glm_pattern_type <- "GLM"
pattern_type <- "Leroux"

# Read in kidney data
seu_kid <- readRDS(paste0(processed_data_path, "seu_kidney.rds"))
seu_kid <- seu_kid[1:3000, ]

# Read in pseudobulk results
limma_df_all <- readRDS(paste0(processed_data_path, "limma_voom_real_markergene_natgen_kidney.rds"))
deseq2_df_all <- readRDS(paste0(processed_data_path, "DESeq2_real_markergene_natgen_kidney.rds"))

# Rename models
limma_df_all <- limma_df_all %>%
  mutate(fit_model = recode(fit_model,
                            "limma_voom" = "limma-voom"))

# Read in GLM model results
glm_wald_df_all <- readRDS(paste0(processed_data_path, "glm_gam_model_results_waldstats_markergene_onlyinteract.rds")) 
# Handle missing columns
glm_wald_df_all$contrast_description <- glm_wald_df_all$contrast_name
if (!("fit_model" %in% colnames(glm_wald_df_all))) {
  glm_wald_df_all$fit_model <- glm_wald_df_all$model
}
# Rename models
glm_wald_df_all <- glm_wald_df_all %>%
  mutate(fit_model = recode(fit_model,
                            "Poisson_GLM" = "Poisson GLM",
                            "NB_GLM" = "NB GLM",
                            "GAM" = "GAM"))

# Add in p-values and adjusted p-values
glm_wald_df_all$wald_pval <- pchisq(glm_wald_df_all$wald_stat_t^2, df = 1, lower.tail = FALSE)
glm_wald_df_all$wald_pval_adj <- NA
for (fm in unique(glm_wald_df_all$fit_model)) {
  for (cd in unique(glm_wald_df_all$contrast_description)) {
    row_idx <- which((glm_wald_df_all$fit_model == fm) & (glm_wald_df_all$contrast_description == cd))
    glm_wald_df_all$wald_pval_adj[row_idx] <- p.adjust(glm_wald_df_all$wald_pval[row_idx], method = "BH")
  }
}

# Read in TESSERA results from fitting all genes and contrasts FOR WITHIN CELL TYPE
# Just to get parameters
wald_df_all <- readRDS(paste0(processed_data_path, "real_hyptest_tessera_between_celltypes.rds"))
wald_df_all <- wald_df_all$wald_df_all
# Drop old, redundant columns
#wald_df_all <- select(wald_df_all, -c("wald_scaling", "wald_shift"))
# Rename models
wald_df_all <- wald_df_all %>%
  mutate(fit_model = recode(fit_model,
                            "Leroux" = "TESSERA-Leroux"))

```

```{r}
# Adjust p-values per contrast to match TESSERA/GLMs/GAM

# Limma
for (cd in unique(limma_df_all$contrast_description)) {
    row_idx <- which(limma_df_all$contrast_description == cd)
    limma_df_all$pval_adj[row_idx] <- p.adjust(limma_df_all$P.Value[row_idx], method = "BH")
}
# DESeq2
for (cd in unique(deseq2_df_all$contrast_description)) {
    row_idx <- which(deseq2_df_all$contrast_description == cd)
    deseq2_df_all$pval_adj[row_idx] <- p.adjust(deseq2_df_all$pvalue[row_idx], method = "BH")
}
```

```{r}
# Colors for models when using color and fill in ggplots
model_color_vals <- c("TESSERA-Leroux" = "#800080", "TESSERA-CAR" = "hotpink4", "TESSERA-SAR" = "darkred", "TESSERA-spNNGP" = "#006400", "DESeq2" = "darkorange", "limma-voom" = "goldenrod1", "Poisson GLM" = "steelblue4", "NB GLM" =  "#00008B", "GAM" = "tan4")
model_fill_vals <- c("TESSERA-Leroux" = "#C77CFF", "TESSERA-CAR" = "pink", "TESSERA-SAR" = "indianred1", "TESSERA-spNNGP" = "#32CD32", "DESeq2" = "darkorange", "limma-voom" = "goldenrod1", "Poisson GLM" = "lightblue", "NB GLM" =  "cornflowerblue", "GAM" = "burlywood")
```

\newpage

# Marker genes

```{r}
df_markergenes <- read.xlsx(paste0(raw_data_path, "2021-07-12390D-Supplementary Table 5.xlsx"), startRow=6, skipEmptyCols = TRUE, sheet = 1)

# Marker genes

df_map_broad <- data.frame(Subclass.Level.1 = c("PT", "TAL", "DCT", "CNT", "PC", "IC", 
                                                "EC", "VSM/P", "FIB", "IMM"),
                           Subclass.Level.3 = c(NA, NA, NA, NA, NA, NA, NA, NA, NA, NA)
                           )
df_map_broad <- merge(df_map_broad, df_markergenes[, c("Subclass.Level.1", "Subclass.Level.3", "Positive.Markers")], by = c("Subclass.Level.1", "Subclass.Level.3"))


df_map <- data.frame(celltype = c("Fib", "CD4T", "MyoFib", "iPT", "VSMC/Pericyte", 
                                  "GS_Stromal", "Mono", "C_TAL", "CNT", "PT_S1", "PC", "PT_S3", 
                                  "IC", "Plasma_Cells", "Endo_G", "PEC", "DCT", "Podo", 
                                  "Dedifferentiated_Tubule", "Mes", "Mac", "Endo_Peritubular", 
                                  "Endo_Lym", "PT_S2", "M_TAL", "RBC"),
                     Subclass.Level.3 = c("FIB", "T", "MYOF", NA, "VSMC/P",
                                          NA, NA, "C-TAL", "CNT", "PT-S1", NA, "PT-S3",
                                          NA, "PL", "EC-GC", "PEC", NA, "POD",
                                          NA, "MC", NA, "EC-PTC",
                                          "EC-LYM", "PT-S2", "M-TAL", NA),
                     Subclass.Level.1 = c("FIB", "IMM", "FIB", NA, "VSM/P",
                                          NA, NA, "TAL", "CNT", "PT", "PC", "PT",
                                          "IC", "IMM", "EC", "PEC", "DCT", "POD", 
                                          NA, "VSM/P", NA, "EC",
                                          "EC", "PT", "TAL", NA)
                     )
df_map <- df_map[!(is.na(df_map$Subclass.Level.1) & is.na(df_map$Subclass.Level.3)), ]

df_map <- merge(df_map, df_markergenes[, c("Subclass.Level.1", "Subclass.Level.3", "Positive.Markers")], by = c("Subclass.Level.1", "Subclass.Level.3"))
df_map <- df_map %>%
  group_by(Subclass.Level.1, Subclass.Level.3, celltype) %>%
  summarise(Positive.Markers = paste(Positive.Markers, collapse = ", "), .groups = "drop")
df_map$Positive.Markers <- gsub(",,", ", ", df_map$Positive.Markers)

df_map_fin <- merge(df_map, df_map_broad[, c("Subclass.Level.1", "Positive.Markers")],
                    by = "Subclass.Level.1", all.x=TRUE)
df_map_fin$Positive.Markers <- paste0(df_map_fin$Positive.Markers.y, ", ", df_map_fin$Positive.Markers.x)
df_map_fin <- df_map_fin[, -which(colnames(df_map_fin) %in% c("Positive.Markers.x", "Positive.Markers.y"))]
write.csv(df_map_fin, paste0(processed_data_path, "marker_genes.csv"), row.names=FALSE)
```

```{r}
paper_mg <- c("EMCN", "PLVAP", "ALB", "PDGFRA", "C7", "COL1A1", "SERPINE1", "ACTA2", "SYNPO2", "MYH11", "ITGB6", "IGFBP6", "CFH", "WT1", "NPHS1", "SLC22A8", "SLC4A4", "SLC7A13", "HAVCR1", "VCAM1", "SLC34A2", "SLC12A1", "SIM2", "SLC12A3", "HSD11B2", "AQP2", "SLC26A7", "ATP6V0D2", "JCHAIN", "CD96", "LYZ", "C1QA", "HBA2")
```

We begin by visualizing all of the cell-types and all marker genes. 
There is some degree of patterning/signal present where we see there are groups of genes that are highly expressed in a particular cell type(s) but not the rest.

```{r, fig.height=15, fig.width=20}
df_map_fin <- read.csv(paste0(processed_data_path, "marker_genes.csv"))
pos_mark <- paste(df_map_fin$Positive.Markers, collapse = ", ")
pos_mark <- trimws(unlist(strsplit(pos_mark, ",")))

seu_obj <- seu_kid[, which(seu_kid@meta.data$celltype %in% df_map_fin$celltype)]
seu_obj$celltype <- factor(seu_obj$celltype, levels = df_map_fin$celltype)
seu_obj <- seu_obj[unique(pos_mark), ]
seu_obj <- Seurat::NormalizeData(seu_obj)

p1 <- Seurat::DotPlot(object = seu_obj,
                      features = unique(pos_mark), group.by = "celltype", scale = TRUE)
p1 <- p1 + labs(x = "Marker Gene", y = "Cell Type")
p1 <- p1 + ggplot2::theme(text = ggplot2::element_text(size = 25))
p1 <- p1 + ggplot2::theme(axis.text.x = ggplot2::element_text(size = 8, angle = 90, vjust = 0.5, hjust=1))
p1 <- p1 + ggplot2::theme(axis.text.y = ggplot2::element_text(size = 20))
p1
ggsave(paste0(figure_path, "marker_genes.pdf"), p1, width = 20, height=16, units = "in", dpi = 300)
```

```{r}
for(c_idx in 1:length(unique(df_map_fin$celltype))){
  mg <- as.character(df_map_fin[which(df_map_fin$celltype == unique(df_map_fin$celltype)[c_idx]), "Positive.Markers"])
  mg <- trimws(unlist(strsplit(mg, ",")))
  mg <- unique(mg)
#  seu_kid@meta.data$celltype <- factor(seu_kid@meta.data$celltype)
#  levels(seu_kid@meta.data$celltype) <- c(setdiff(levels(seu_kid@meta.data$celltype), unique(df_map_fin$celltype)[c_idx]), unique(df_map_fin$celltype)[c_idx])
  
  p1 <- Seurat::DotPlot(object = seu_kid[, which(seu_kid@meta.data$celltype %in% unique(df_map_fin$celltype))], features = mg[mg %in% rownames(seu_kid)], group.by = "celltype")
  p1 <- p1 + ggplot2::theme(text = ggplot2::element_text(size = 20))
  p1 <- p1 + ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90, vjust = 0.5, hjust=1))
  ggsave(paste0(figure_path, "indiv_ctyp_mg/marker_genes_", gsub("/", "_", unique(df_map_fin$celltype)[c_idx]), ".pdf"), p1, width = 12, height=9, units = "in", dpi = 300)
}
```

```{r, fig.height=8, fig.width=10}
pt_mg <- as.character(df_map_fin[which(df_map_fin$celltype == "PT_S1"), "Positive.Markers"])
pt_mg <- unique(trimws(unlist(strsplit(pt_mg, ","))))

ctal_mg <- as.character(df_map_fin[which(df_map_fin$celltype == "C_TAL"), "Positive.Markers"])
ctal_mg <- unique(trimws(unlist(strsplit(ctal_mg, ","))))

mg <- c(pt_mg, ctal_mg)

p1 <- Seurat::DotPlot(object = seu_kid[, which(seu_kid@meta.data$celltype %in% unique(df_map_fin$celltype))], features = mg[mg %in% rownames(seu_kid)], group.by = "celltype")
p1 <- p1 + geom_rect(aes(xmin = 0, xmax = length(pt_mg) + 0.5, ymin = 16.5, ymax = 17.5), color = "red", fill = NA, size = 1)
p1 <- p1 + geom_rect(aes(xmin = length(pt_mg) + 0.5, xmax = length(mg) + 0.5, ymin = 0.5, ymax = 1.5), color = "red", fill = NA, size = 1)
p1 <- p1 + labs(x = "Marker Gene", y = "Cell Type")
p1 <- p1 + ggplot2::theme(text = ggplot2::element_text(size = 20))
p1 <- p1 + ggplot2::theme(axis.text.x = ggplot2::element_text(size = 20, angle = 90,
                                                              vjust = 0.5, hjust=1))
p1 <- p1 + ggplot2::theme(axis.text.y = ggplot2::element_text(size = 20))
p1
ggsave(paste0(figure_path, "marker_genes_ctal_pt_allctyp.pdf"), p1, width = 12, height=9, units = "in", dpi = 300)
```

\newpage

We next subset to just the two cell-types of interest, PT_S1 and C_TAL. 
Here, we can see that for most of the PT_S1 marker genes, we should hope to find a signal, and the same for the C_TAL genes. 

```{r}
pt_mg <- as.character(df_map_fin[which(df_map_fin$celltype == "PT_S1"), "Positive.Markers"])
pt_mg <- unique(trimws(unlist(strsplit(pt_mg, ","))))

ctal_mg <- as.character(df_map_fin[which(df_map_fin$celltype == "C_TAL"), "Positive.Markers"])
ctal_mg <- unique(trimws(unlist(strsplit(ctal_mg, ","))))

mg <- c(pt_mg, ctal_mg)
mg <- mg[mg %in% rownames(seu_kid)]
mg_all <- mg

seu_obj <- seu_kid[, which(seu_kid@meta.data$celltype %in% c("PT_S1", "C_TAL"))]
seu_obj <- seu_obj[mg[mg %in% rownames(seu_kid)], ]
seu_obj <- seu_obj[unique(pos_mark), ]
seu_obj <- Seurat::NormalizeData(seu_obj)

p1 <- Seurat::DotPlot(object = seu_obj, features = mg, group.by = "celltype", scale=TRUE)
p1 <- p1 + labs(x = "Marker Gene", y = "Cell Type")
p1 <- p1 + geom_rect(aes(xmin = 0, xmax = length(pt_mg) + 0.5, ymin = 1.5, ymax = 2.5), color = "red", fill = NA, size = 1)
p1 <- p1 + geom_rect(aes(xmin = length(pt_mg) + 0.5, xmax = length(mg) + 0.5, ymin = 0.5, ymax = 1.5), color = "red", fill = NA, size = 1)
p1 <- p1 + ggplot2::theme(text = ggplot2::element_text(size = 15))
p1 <- p1 + ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90,
                                                              vjust = 0.5, hjust=1))
p1
ggsave(paste0(figure_path, "marker_genes_ctal_pt.pdf"), p1, width = 8, height=6, units = "in", dpi = 300)
```

```{r}
# Ranks
wald_df_all$gene_rank <- NA
for (mv in unique(wald_df_all$fit_model)) {
  for (cd in unique(wald_df_all$contrast_description)) {
    tmp_idx <- which((wald_df_all$fit_model == mv) & (wald_df_all$contrast_description == cd))
    wald_df_all$gene_rank[tmp_idx] <- length(tmp_idx) - rank(wald_df_all$wald_stat_t[tmp_idx]^2) + 1
  }
}

glm_wald_df_all$gene_rank <- NA
for (mv in unique(glm_wald_df_all$fit_model)) {
  for (cd in unique(glm_wald_df_all$contrast_description)) {
    tmp_idx <- which((glm_wald_df_all$fit_model == mv) & (glm_wald_df_all$contrast_description == cd))
    glm_wald_df_all$gene_rank[tmp_idx] <- length(tmp_idx) - rank(glm_wald_df_all$wald_stat_t[tmp_idx]^2) + 1
  }
}

limma_df_all$gene_rank <- NA
for (mv in unique(limma_df_all$fit_model)) {
  for (cd in unique(limma_df_all$contrast_description)) {
    tmp_idx <- which((limma_df_all$fit_model == mv) & (limma_df_all$contrast_description == cd))
    limma_df_all$gene_rank[tmp_idx] <- length(tmp_idx) - rank(abs(limma_df_all$t[tmp_idx])) + 1
  }
}

deseq2_df_all$gene_rank <- NA
for (mv in unique(deseq2_df_all$fit_model)) {
  for (cd in unique(deseq2_df_all$contrast_description)) {
    tmp_idx <- which((deseq2_df_all$fit_model == mv) & (deseq2_df_all$contrast_description == cd))
    deseq2_df_all$gene_rank[tmp_idx] <- length(tmp_idx) - rank(abs(deseq2_df_all$stat[tmp_idx])) + 1
  }
}
```

```{r}
# Count number of significant genes per contrast
PVAL_THRESH <- 0.05
id_cols <- c("fit_model", 
             "kernel_type", 
             #"contrast_string", 
             "contrast_indices", 
             "contrast_description"
             )
wald_gene_nsig_df <- data.frame(
  wald_df_all
  %>% dplyr::group_by(dplyr::across(dplyr::all_of(id_cols)))
  %>% dplyr::summarise(n_sig_genes = sum(wald_pval_adj < PVAL_THRESH, na.rm =
                                           TRUE))
)
```

```{r, fig.height=8, fig.width=10}
# p <- ggplot2::ggplot(wald_gene_nsig_df, ggplot2::aes(x = contrast_description, 
#                                              y = n_sig_genes,
#                                              fill = contrast_description
#                                              ))
# p <- p + ggplot2::geom_col(position = "dodge", color = "black")
# p <- p + ggplot2::scale_fill_manual(values = c("cornflowerblue", "goldenrod1", "lightcoral", "forestgreen", "gray", "darkorchid"))
# p <- p + ggplot2::labs(x = "Contrast",
#                        y = "Number of significant genes\n(p.adj < 0.05)",
#                        fill = "Comparison"
#                        )
# p <- p + ggplot2::theme_bw()
# p <- p + ggplot2::theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust=0.5))
# p <- p + ggplot2::theme(text = element_text(size=20))
# p
```

```{r}
# Count number of significant genes per contrast
PVAL_THRESH <- 0.05
glm_id_cols <- c("fit_model", 
             # "contrast_string", 
             "contrast_indices", 
             "contrast_description"
             )
glm_gene_nsig_df <- data.frame(
  glm_wald_df_all
  %>% dplyr::group_by(dplyr::across(dplyr::all_of(glm_id_cols)))
  %>% dplyr::summarise(n_sig_genes = sum(wald_pval_adj < PVAL_THRESH, na.rm =
                                           TRUE))
)

# Count number of significant genes per contrast
PVAL_THRESH <- 0.05
limma_id_cols <- c("fit_model", 
             # "contrast_string", 
             # "contrast_indices", 
             "contrast_description"
             )
limma_gene_nsig_df <- data.frame(
  limma_df_all
  %>% dplyr::group_by(dplyr::across(dplyr::all_of(limma_id_cols)))
  %>% dplyr::summarise(n_sig_genes = sum(pval_adj < PVAL_THRESH, na.rm =
                                           TRUE))
)

# Count number of significant genes per contrast
PVAL_THRESH <- 0.05
deseq2_id_cols <- c("fit_model", 
             # "contrast_string", 
             # "contrast_indices", 
             "contrast_description"
             )
deseq2_gene_nsig_df <- data.frame(
  deseq2_df_all
  %>% dplyr::group_by(dplyr::across(dplyr::all_of(deseq2_id_cols)))
  %>% dplyr::summarise(n_sig_genes = sum(pval_adj < PVAL_THRESH, na.rm =
                                           TRUE))
)

plot_data_sub <- wald_gene_nsig_df[, c("fit_model", "contrast_description", "n_sig_genes")]
glm_plot_data_sub <- glm_gene_nsig_df[, c("fit_model", "contrast_description", "n_sig_genes")]
limma_data_sub <- limma_gene_nsig_df[, c("fit_model", "contrast_description", "n_sig_genes")]
deseq2_data_sub <- deseq2_gene_nsig_df[, c("fit_model", "contrast_description", "n_sig_genes")]
plot_data_comb <- rbind(plot_data_sub, glm_plot_data_sub, limma_data_sub, deseq2_data_sub)
plot_data_comb$fit_model <- factor(plot_data_comb$fit_model, levels=c("TESSERA-Leroux", "DESeq2", "limma-voom", "Poisson GLM", "NB GLM", "GAM"))
```

```{r, fig.height=8, fig.width=10}
p <- ggplot2::ggplot(plot_data_comb, ggplot2::aes(x = fit_model, 
                                             y = n_sig_genes,
                                             fill = fit_model
                                             ))
p <- p + ggplot2::geom_col(position = "dodge", color = "black")
p <- p + ggplot2::facet_wrap(~contrast_description)
p <- p + ggplot2::scale_fill_manual(values = model_fill_vals)
p <- p + ggplot2::labs(x = "Contrast",
                       y = "Number of significant genes\n(p.adj < 0.05)",
                       fill = "Comparison"
                       )
p <- p + ggplot2::theme_bw()
p <- p + ggplot2::theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust=0.5))
p <- p + ggplot2::theme(text = element_text(size=20))
p
```

```{r, fig.height=8, fig.width=10}
p <- ggplot2::ggplot(plot_data_comb[which(plot_data_comb$contrast_description == "C_TAL v. PT_S1"), ], ggplot2::aes(x = fit_model, y = n_sig_genes, fill = fit_model))
p <- p + ggplot2::geom_col(position = "dodge", color = "black")
p <- p + ggplot2::facet_wrap(~contrast_description)
p <- p + ggplot2::scale_fill_manual(values = model_fill_vals, name = "Fitted Model")
p <- p + ggplot2::labs(x = "Fitted Model",
                       y = "Number of significant genes (FDR < 0.05)"
                       )
p <- p + ggplot2::theme_bw()
p <- p + ggplot2::theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust=0.5))
p <- p + ggplot2::theme(text = element_text(size=20), legend.position = "none")
p
ggsave(paste0(figure_path, "num_sig_genes_ctal_pt.pdf"), p, width = 8, height=8, units = "in")
```

```{r}
# Subset to comparison of interest: C_TAL v. PT_S1
limma_df_all <- limma_df_all[which(limma_df_all$contrast_description == "C_TAL v. PT_S1"), ]
deseq2_df_all <- deseq2_df_all[which(deseq2_df_all$contrast_description == "C_TAL v. PT_S1"), ]
glm_wald_df_all <- glm_wald_df_all[which(glm_wald_df_all$contrast_description == "C_TAL v. PT_S1"), ]
wald_df_all <- wald_df_all[which(wald_df_all$contrast_description == "C_TAL v. PT_S1"), ]
```

```{r}
# Jaccard similarity of top-n genes for each pair of methods and contrast
tmp_wald_df <- dplyr::bind_rows(limma_df_all, deseq2_df_all, glm_wald_df_all, wald_df_all)
n_list <- seq(1, 3000)

plot_df <- list()
for (mv1 in unique(tmp_wald_df$fit_model)) { # Loop over method 1
  for (mv2 in unique(tmp_wald_df$fit_model)) { # Loop over method 2
    if (mv1 == mv2) { # Don't bother with the same same method (trivial)
      next
    } else if (mv2 < mv1) { # We don't need to compare both (x, y) and (y, x)
      next
    }
    
    for (cd in unique(tmp_wald_df$contrast_description)) { # Loop over contrasts
      #cat(Sys.time(), mv1, mv2, cd, "\n")
      for (idx in 1:length(n_list)) {
        g1 <- tmp_wald_df$gene[(tmp_wald_df$gene_rank <= n_list[idx]) & (tmp_wald_df$fit_model == mv1)]
        g2 <- tmp_wald_df$gene[(tmp_wald_df$gene_rank <= n_list[idx]) & (tmp_wald_df$fit_model == mv2)]
        
        plot_df[[1 + length(plot_df)]] <- data.frame(
          fit_model1 = mv1,
          fit_model2 = mv2,
          contrast_description = cd,
          n = n_list[idx],
          n_intersect = length(intersect(g1, g2)),
          n_union = length(unique(c(g1, g2)))
        )
      }
    }
  }
}
plot_df <- dplyr::bind_rows(plot_df)
plot_df$jaccard <- plot_df$n_intersect / plot_df$n_union
```

```{r}
# Expected Jaccard similarity of random ordering
j_trials <- 100
j_mat <- array(data=NA, dim=c(length(n_list), j_trials, 2))
gene_list <- unique(tmp_wald_df$gene)
for (tr in 1:j_trials) {
  g1 <- sample(gene_list, replace = FALSE)
  g2 <- sample(gene_list, replace = FALSE)

  for (idx in 1:nrow(j_mat)) {
    j_mat[idx, tr, 1] <- length(intersect(g1[1:n_list[idx]], g2[1:n_list[idx]]))
    j_mat[idx, tr, 2] <- length(union(g1[1:n_list[idx]], g2[1:n_list[idx]]))
  }
}
exp_jaccard_data <- data.frame(n=n_list, expected_jaccard=rowMeans(j_mat[, , 1] / j_mat[, , 2]))

plot_df <- merge(plot_df, exp_jaccard_data, by="n")
```

\newpage

We compare the ranking of the genes by their Wald statistics for all models. 
That is, for each pair of models (the GLMs and the Poisson-Leroux), as a function of the number of top genes $n$, we compute the Jaccard similarity (intersection divided by union) of the top-$n$ genes found by each method. 
We compare these curves to the expected Jaccard similarity, assuming that the rankings were uniform and random. 
To calculate the expected Jaccard similarity, for each value of n, randomly draw two lists of $n$ genes.
Next, compute the Jaccard similarity between those two lists and repeat this procedure for 100 trials. Finally, take the average Jaccard similarity across all trials.

```{r, fig.height=8, fig.width=10}
color_vals <- c("#d62728","#1f77b4","#2ca02c","#ff7f0e","#9467bd","#8c564b","#e377c2",
                "#7f7f7f","#bcbd22","#17becf","#ff9896","#aec7e8","#98df8a","#ffbb78", "brown")
plot_df$method_comparison <- paste0(plot_df$fit_model1, " v. ", plot_df$fit_model2)
p1 <- ggplot2::ggplot(data=plot_df, ggplot2::aes(x=n, 
                                                 y=jaccard,
                                                 color=method_comparison
                                                 ))
p1 <- p1 + ggplot2::geom_point(size=1)
p1 <- p1 + ggplot2::geom_line(ggplot2::aes(x=n, y=expected_jaccard), color="black", lwd=1.5)
p1 <- p1 + scale_color_manual(values = color_vals) #c("goldenrod", "lightcoral", "cornflowerblue", "purple", "seagreen", "darkorange"
p1 <- p1 + ggplot2::guides(color = guide_legend(override.aes = list(size = 3)))
p1 <- p1 + ggplot2::labs(x = expression("Top " * italic(n) * " Genes"), 
                         y = "Jaccard Similarity",
                         color = "Comparison"
                         )
p1 <- p1 + ggplot2::theme_bw()
p1 <- p1 + ggplot2::theme(text=ggplot2::element_text(size=20),
                          axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
p1
ggsave(paste0(figure_path, "jaccard_all.pdf"), p1, width = 8, height=6, units = "in")
```
```{r, fig.height=8, fig.width=10}
plot_df_sub_med <- plot_df[which(plot_df$method_comparison %in% c("DESeq2 v. limma-voom", "NB GLM v. Poisson GLM", "GAM v. Poisson GLM", "GAM v. NB GLM", "limma-voom v. TESSERA-Leroux", "DESeq2 v. TESSERA-Leroux", "Poisson GLM v. TESSERA-Leroux", "NB GLM v. TESSERA-Leroux", "GAM v. TESSERA-Leroux")), ]
plot_df_sub_med$method_comparison <- factor(plot_df_sub_med$method_comparison, c("DESeq2 v. limma-voom", "NB GLM v. Poisson GLM", "GAM v. Poisson GLM", "GAM v. NB GLM", "limma-voom v. TESSERA-Leroux", "DESeq2 v. TESSERA-Leroux", "Poisson GLM v. TESSERA-Leroux", "NB GLM v. TESSERA-Leroux", "GAM v. TESSERA-Leroux"))

p1 <- ggplot2::ggplot(data=plot_df_sub_med, ggplot2::aes(x=n, 
                                                 y=jaccard,
                                                 color=method_comparison
                                                 ))
p1 <- p1 + ggplot2::geom_point(size=1)
p1 <- p1 + ggplot2::geom_line(ggplot2::aes(x=n, y=expected_jaccard), color="black", lwd=1.5)
p1 <- p1 + facet_wrap(~contrast_description)
p1 <- p1 + scale_color_manual(values = c("#d62728", "purple", "navy", "cornflowerblue", "seagreen", "lightgreen", "sienna", "darkorange", "goldenrod3", "darkgray"))
p1 <- p1 + ggplot2::labs(x = expression("Top " * italic(n) * " Genes"), 
                         y = "Jaccard Similarity",
                         color = "Comparison"
                         )
p1 <- p1 + ggplot2::theme_bw()
p1 <- p1 + ggplot2::theme(text = ggplot2::element_text(size = 20),
                          axis.text.x = ggplot2::element_text(angle = 90, vjust = 0.5, hjust = 1),
                          legend.text = ggplot2::element_text(size = 12),
                          legend.title = ggplot2::element_text(size = 16))
p1 <- p1 + ggplot2::guides(color = guide_legend(override.aes = list(size = 3)))
p1
ggsave(paste0(figure_path, "jaccard.pdf"), p1, width = 8, height=6, units = "in")
```

```{r, fig.height=8, fig.width=10}
plot_df_sub <- plot_df[which(plot_df$method_comparison %in% c("DESeq2 v. limma-voom", "DESeq2 v. TESSERA-Leroux", "limma-voom v. TESSERA-Leroux")), ]
plot_df_sub$method_comparison <- factor(plot_df_sub$method_comparison, levels = c("DESeq2 v. limma-voom", "limma-voom v. TESSERA-Leroux", "DESeq2 v. TESSERA-Leroux"))

p1 <- ggplot2::ggplot(data=plot_df_sub, ggplot2::aes(x=n, 
                                                 y=jaccard,
                                                 color=method_comparison
                                                 ))
p1 <- p1 + ggplot2::geom_point(size=1)
p1 <- p1 + ggplot2::geom_line(ggplot2::aes(x=n, y=expected_jaccard), color="black", lwd=1.5)
p1 <- p1 + scale_color_manual(values = c("cornflowerblue", "lightcoral", "goldenrod"))
p1 <- p1 + ggplot2::guides(color = guide_legend(override.aes = list(size = 3)))
p1 <- p1 + ggplot2::labs(x = expression("Top " * italic(n) * " Genes"), 
                         y = "Jaccard Similarity",
                         color = "Comparison"
                         )
p1 <- p1 + ggplot2::theme_bw()
p1 <- p1 + ggplot2::theme(text=ggplot2::element_text(size=20),
                          axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
p1
ggsave(paste0(figure_path, "jaccard_pb.pdf"), p1, width = 8, height=6, units = "in")
```

\newpage



```{r, fig.height=8, fig.width=10}
p <- ggplot2::ggplot(plot_data_comb, ggplot2::aes(x = fit_model, 
                                             y = n_sig_genes,
                                             fill = fit_model
                                             ))
p <- p + ggplot2::geom_col(position = "dodge")
p <- p + ggplot2::scale_fill_manual(values = model_fill_vals)
p <- p + ggplot2::labs(x = "Contrast",
                       y = "Number of significant genes\n(p.adj < 0.05)",
                       fill = "Comparison"
                       )
p <- p + ggplot2::theme_bw()
p <- p + ggplot2::theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust=0.5))
p <- p + ggplot2::theme(text = element_text(size=20))
p
```

We next plot the ranks of the marker genes in the top 3000 genes by variance, as ranked by the magnitude of the Wald statistics for each method.
We see that, in general, both the TESSERA Poisson-Leroux model and the GLMs rank the marker genes highly. 
We also note that the rankings are similar across conditions, as expected.

```{r, fig.height=8, fig.width=10}
rank_plot <- tmp_wald_df
rank_plot <- rank_plot[which(rank_plot$gene %in% mg_all), ]
rank_plot$fit_model <- factor(rank_plot$fit_model, levels = c("TESSERA-Leroux", "DESeq2", "limma-voom", "Poisson GLM", "NB GLM", "GAM"))
rank_plot$gene <- factor(rank_plot$gene, levels = mg_all)
p1 <- ggplot2::ggplot(data=rank_plot, ggplot2::aes(x=gene, y=gene_rank, fill=fit_model))
p1 <- p1 + ggplot2::geom_point(size=4, shape = 21, stroke = 0.5, color = "black") 
p1 <- p1 + ggplot2::scale_fill_manual(values=model_fill_vals, name = "Fitted Model")
p1 <- p1 + ggplot2::facet_wrap(~contrast_description)
p1 <- p1 + ggplot2::theme_bw()
p1 <- p1 + ggplot2::labs(x = "Gene", y = "Rank")
p1 <- p1 + ggplot2::scale_y_continuous(trans = "sqrt", breaks = c(0, 50, 100, 250, 500, 1000, 2000, 3000))
p1 <- p1 + ggplot2::theme(text=ggplot2::element_text(size=20),
                          axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
                          panel.grid.minor = element_blank())
p1
ggsave(paste0(figure_path, "ranks_ctal_pt.pdf"), p1, width = 8, height=6, units = "in")
```

\newpage

We present a volcano plot to compare the Benjamini-Hochberg (BH) adjusted p-values (note, not the Wald statistics) with the contrast values.
We perform a Benjamini-Hochberg adjustment for each set of p-values associated to the 3000 genes for a particular condition and model type.
Both the Poisson-Leroux method and the GLMs have large beta (contrast) values for the marker genes and most also have small p-values, as you would expect.
We note that there is a noticeable gap in the Poisson-Leroux volcano plots as we compute conditional p-values, meaning that any Wald statistic above our determined threshold gets ruled as significant and corresponds to a p-value of 0.

```{r, fig.height=15, fig.width=20}
# Volcano plot for TESSERA
volc_plot <- tmp_wald_df
volc_plot$marker_gene <- FALSE
volc_plot[which(volc_plot$gene %in% mg), "marker_gene"] <- TRUE

volc_plot$pval_adj_all <- volc_plot$wald_pval_adj
pb_idx <- volc_plot$fit_model %in% c("DESeq2", "limma-voom")
volc_plot$pval_adj_all[pb_idx] <- volc_plot$pval_adj[pb_idx]

volc_plot$contrast_val_all <- volc_plot$contrast_val
volc_plot$contrast_val_all[which(volc_plot$fit_model == "DESeq2")] <- volc_plot$log2FoldChange[which(volc_plot$fit_model == "DESeq2")]
volc_plot$contrast_val_all[which(volc_plot$fit_model == "limma-voom")] <- volc_plot$logFC[which(volc_plot$fit_model == "limma-voom")]


# So that we can plot zero p-values
volc_plot$wald_pval_adj[volc_plot$wald_pval_adj < .Machine$double.eps] <- .Machine$double.eps
volc_plot$pval_adj_all[volc_plot$pval_adj_all < .Machine$double.eps] <- .Machine$double.eps

volc_plot$fit_model <- factor(volc_plot$fit_model, levels=c("TESSERA-Leroux", "DESeq2", "limma-voom", "Poisson GLM", "NB GLM", "GAM"))

p1 <- ggplot2::ggplot(data=volc_plot, ggplot2::aes(x=contrast_val_all,
                                                y=-log10(pval_adj_all),
                                                color=marker_gene))
p1 <- p1 + ggplot2::facet_wrap(~fit_model, ncol=2, scales = "free", labeller = 
      labeller(.multi_line=FALSE))
p1 <- p1 + ggplot2::geom_point(size=2, alpha=0.5)
p1 <- p1 + ggplot2::scale_color_manual(values=c("darkgray", "red"), name = "Marker Gene")
p1 <- p1 + ggplot2::coord_cartesian(clip = 'off')
p1 <- p1 + ggplot2::theme_bw()
p1 <- p1 + ggplot2::labs(x = "Estimated Contrast Value", y = "-log10(p-value)")
p1 <- p1 + ggplot2::theme(text=ggplot2::element_text(size=20),
                          axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
p1
ggsave(paste0(figure_path, "volcano_plot.pdf"), p1, width = 12, height=12, units = "in")
```
