---
title: "Real data results"
subtitle: "Differentially expressed genes between conditions within a cell type"
author: "Florica"
date: "2025-09-01"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE,
                      message = FALSE)
```

```{r}
# Library
library(dplyr)
library(forcats)
library(purrr)
library(ggplot2)
library(patchwork)

# Paths
processed_data_path <- "../../data/processed/"
raw_data_path <- "../../data/raw/"
figure_path <- "../../figures/real/DE_between_conditions_within_celltype/"

# Model type
glm_pattern_type <- "GLM"
pattern_type <- "Leroux"

# Read in kidney data
seu_kid <- readRDS(paste0(processed_data_path, "seu_kidney.rds"))
seu_kid@meta.data$sample <- seu_kid@meta.data$orig.ident

# Read in pseudobulk results
limma_df_all <-  readRDS(paste0(processed_data_path, "limma_voom_real_natgen_kidney.rds"))
deseq2_df_all <-  readRDS(paste0(processed_data_path, "DESeq2_real_natgen_kidney.rds"))

# Read in GLM model results
glm_wald_df_all <- readRDS(paste0(processed_data_path, "glm_gam_model_results_waldstats_onlyinteract.rds"))
glm_summary_out_all <- readRDS(paste0(processed_data_path, "glm_gam_model_results_perfsummary_onlyinteract.rds"))

wald_df_all <- readRDS(paste0(processed_data_path, "real_hyptest_tessera.rds"))
wald_thresh_df <- wald_df_all$fit_results_df
wald_pval_qq_df <- wald_df_all$pval_qq_df
wald_df_all <- wald_df_all$wald_df_all

summary_out_all <- readRDS(paste0(processed_data_path, "real_results_perfsummary_", pattern_type, "_onlyinteract.rds"))
# Add in diagnosis information
summary_out_all <- dplyr::left_join(summary_out_all, dplyr::distinct(seu_kid@meta.data[, c("sample", "Group")]), by = "sample")
summary_out_all$Group <- factor(summary_out_all$Group, levels = c("Control", "HKD", "DKD"))
summary_out_all$sample <- forcats::fct_reorder(summary_out_all$sample, as.numeric(summary_out_all$Group))

# Rename models
limma_df_all <- limma_df_all %>%
  mutate(fit_model = recode(fit_model,
                            "limma_voom" = "limma-voom"))

wald_df_all <- wald_df_all %>%
  mutate(fit_model = recode(fit_model,
                            "Leroux" = "TESSERA-Leroux"))

glm_wald_df_all <- glm_wald_df_all %>%
  mutate(fit_model = recode(fit_model,
                            "NB_GLM" = "NB GLM",
                            "Poisson_GLM" = "Poisson GLM"))

summary_out_all <- summary_out_all %>%
  mutate(fit_model = recode(fit_model,
                            "Leroux" = "TESSERA-Leroux"))

glm_summary_out_all <- glm_summary_out_all %>%
  mutate(fit_model = recode(fit_model,
                            "Leroux" = "TESSERA-Leroux"))
```

```{r}
# Adjust p-values per contrast to match TESSERA/GLMs/GAM

# Limma
for (cd in unique(limma_df_all$contrast_description)) {
    row_idx <- which(limma_df_all$contrast_description == cd)
    limma_df_all$pval_adj[row_idx] <- p.adjust(limma_df_all$P.Value[row_idx], method = "BH")
}
# DESeq2
for (cd in unique(deseq2_df_all$contrast_description)) {
    row_idx <- which(deseq2_df_all$contrast_description == cd)
    deseq2_df_all$pval_adj[row_idx] <- p.adjust(deseq2_df_all$pvalue[row_idx], method = "BH")
}
```

# Hypothesis Testing


```{r, fig.height=4, fig.width=6}
tmp_plot <- wald_thresh_df
tmp_plot$Raw_RMSE <- sqrt(tmp_plot$Raw_MSE)
tmp_plot <- reshape2::melt(tmp_plot, c("quantile", "threshold")) %>%
                       dplyr::filter(grepl("Raw_", variable))
tmp_plot$variable <- gsub("Raw_", "", tmp_plot$variable)
tmp_plot <- tmp_plot[tmp_plot$variable != "MSE", ]

p <- ggplot2::ggplot(tmp_plot,
                     ggplot2::aes(x=as.numeric(quantile), 
                                  y=as.numeric(value), 
                                  group=variable,
                                  color=variable
                                  )
                       )
p <- p + ggplot2::geom_line(linewidth = 1)
p <- p + ggplot2::scale_y_log10() #breaks=c(1e-2, 3e-2, 1e-1, 3e-1))
p <- p + ggplot2::labs(x="Quantile", 
                       y="Error",
                       color="Metric"
                       )
p <- p + ggplot2::theme_bw()
p <- p + ggplot2::theme(text = element_text(size=20))
p
ggsave(paste0(figure_path, "errors_metrics_by_quantile.pdf"), p, width = 8, height=6, units = "in", dpi = 300)
```

```{r, fig.height=4, fig.width=6}
tmp_plot <- wald_thresh_df
tmp_plot$Raw_RMSE <- sqrt(tmp_plot$Raw_MSE)
tmp_plot <- reshape2::melt(tmp_plot, c("quantile", "threshold")) %>%
                       dplyr::filter(grepl("Raw_", variable))
tmp_plot$variable <- gsub("Raw_", "", tmp_plot$variable)
tmp_plot <- tmp_plot[tmp_plot$variable != "MSE", ]

p <- ggplot2::ggplot(tmp_plot,
                     ggplot2::aes(x=as.numeric(threshold), 
                                  y=as.numeric(value), 
                                  group=variable,
                                  color=variable
                                  )
                       )
p <- p + ggplot2::geom_line(linewidth = 1)
p <- p + ggplot2::scale_y_log10() #breaks=c(1e-2, 3e-2, 1e-1, 3e-1))
p <- p + ggplot2::labs(x="Threshold", 
                       y="Error",
                       color="Metric"
                       )
p <- p + ggplot2::theme_bw()
p <- p + ggplot2::theme(text = element_text(size=20))
p
ggsave(paste0(figure_path, "errors_metrics_by_threshold.pdf"), p, width = 8, height=6, units = "in", dpi = 300)
```

\newpage


```{r, fig.height=4, fig.width=6}
p <- ggplot2::ggplot(wald_pval_qq_df, aes(x = theoretical, y = observed))
p <- p + ggplot2::geom_line(linewidth = 1)
p <- p + ggplot2::geom_abline(slope = 1, intercept = 0, color="red", linetype="dashed", linewidth = 1)
p <- p + ggplot2::labs(#title="Wald Statistic P-Value v. Uniform Distribution", 
                       #subtitle=paste0("Uniform Fit MSE: ", round(wald_null_MSE, digits = 6)),
                       x="Uniform[0, 1]", y="Estimated P-Values")
p <- p + ggplot2::theme_bw()
p <- p + ggplot2::theme(text = element_text(size=20))
p
ggsave(paste0(figure_path, "QQ_plot_using_wald_threshold.pdf"), p, width = 8, height=6, units = "in", dpi = 300)
```

```{r, fig.height=4, fig.width=6}
p <- ggplot2::ggplot(wald_thresh_df,
                     ggplot2::aes(y=as.numeric(threshold), 
                                  x=as.numeric(quantile))
                       )
p <- p + ggplot2::geom_line(linewidth = 1)
p <- p + ggplot2::geom_hline(yintercept = unique(wald_df_all$wald_thresh)[1], color = "red", linetype = "dashed", linewidth = 1)
p <- p + ggplot2::scale_y_log10()
p <- p + ggplot2::theme_bw()
p <- p + ggplot2::labs(y="Threshold", 
                       x="Quantile"
                       )
p <- p + ggplot2::theme(text = element_text(size=20))
p
ggsave(paste0(figure_path, "quantile_v_wald_threshold.pdf"), p, width = 8, height=6, units = "in", dpi = 300)
```


```{r, fig.height=4, fig.width=7}
tmp_plot <- reshape2::melt(wald_thresh_df[, c("threshold", "scale", "shift", "quantile")], c("quantile", "threshold")) %>%
                       dplyr::filter(variable %in% c("scale", "shift"))
tmp_plot$variable <- stringr::str_to_title(tmp_plot$variable)

p <- ggplot2::ggplot(tmp_plot,
                     ggplot2::aes(x=as.numeric(threshold), y=as.numeric(value), group=variable)
                       )
p <- p + ggplot2::facet_wrap(~variable, scales = "free_y")
p <- p + ggplot2::geom_line(linewidth = 1)
p <- p + ggplot2::geom_vline(xintercept = unique(wald_df_all$wald_thresh)[1], color = "red", linetype = "dashed", linewidth = 1)
p <- p + ggplot2::scale_x_log10()
p <- p + ggh4x::scale_y_facet(
    variable == "Scale",
    trans  = "log10",
    breaks = scales::breaks_log(),
    labels = scales::label_log()
  )
p <- p + ggplot2::labs(x="Threshold", 
                       y="Value"
                       )
p <- p + ggplot2::theme_bw()
p <- p + ggplot2::theme(text = element_text(size=20))
p
ggsave(paste0(figure_path, "scale_and_shift_params.pdf"), p, width = 8, height=6, units = "in", dpi = 300)
```

```{r, fig.height=4, fig.width=6} 
# Create ECDF plot
p <- ggplot2::ggplot(wald_df_all, aes(x = wald_stat_t^2))
p <- p + ggplot2::stat_ecdf(geom = "step", color = "black", linewidth = 1)
p <- p + ggplot2::geom_vline(xintercept = unique(wald_df_all$wald_thresh)[1], color = "red", linewidth = 1, linetype = "dashed")
p <- p + ggplot2::labs(x = "Wald Statistic", y = "ECDF")
p <- p + ggplot2::scale_x_continuous(trans="log10")
p <- p + ggplot2::theme_bw()
p <- p + ggplot2::theme(text = element_text(size=20))
p
ggsave(paste0(figure_path, "ecdf_wald_stats.pdf"), p, width = 8, height=6, units = "in", dpi = 300)
```

```{r, eval = FALSE, fig.height=25, fig.width=25}
wald_df_all$contrast_celltype <- as.vector(sapply(wald_df_all$contrast_description, 
                                                  function (x) { strsplit(x, ":")[[1]][1] }))
wald_df_all$contrast_diseases <- as.vector(sapply(wald_df_all$contrast_description, 
                                                  function (x) { gsub("-", " v. ", 
                                                                      strsplit(x, ":")[[1]][2]) }))


# PLOT: Wald statistics, faceted by cell-type
p <- ggplot2::ggplot(wald_df_all, ggplot2::aes(x = wald_stat_t^2, color = contrast_diseases))
p <- p + ggplot2::stat_ecdf(geom = "step", linewidth = 1)
p <- p + ggplot2::stat_ecdf(geom = "step", linewidth = 1, 
                            ggplot2::aes(x = wald_stat_t^2), color = "black")
p <- p + ggplot2::geom_hline(yintercept = mean(wald_df_all$wald_stat_t^2 < unique(wald_df_all$wald_thresh)[1]), color = "blue", linetype = "dashed")
p <- p + ggplot2::geom_vline(xintercept = unique(wald_df_all$wald_thresh)[1], color = "blue", linetype = "dashed")
p <- p + ggplot2::facet_wrap(~contrast_celltype)
p <- p + ggplot2::labs(x = "Wald F-statistic", y = "ECDF", color = "Comparison")
p <- p + ggplot2::scale_x_continuous(trans="log10")
p <- p + ggplot2::coord_cartesian(xlim = c(1e-3, 1e5))
p <- p + ggplot2::theme_bw()
p <- p + ggplot2::theme(text = element_text(size=20))
p
```

```{r, eval = FALSE, fig.height=25, fig.width=25}
# PLOT: Wald statistics, faceted by disease comparison

p <- ggplot2::ggplot(wald_df_all, ggplot2::aes(x = wald_stat_t^2, color = contrast_celltype))
p <- p + ggplot2::stat_ecdf(geom = "step", linewidth = 1)
p <- p + ggplot2::stat_ecdf(geom = "step", linewidth = 1, 
                            ggplot2::aes(x = wald_stat_t^2), color = "black")
p <- p + ggplot2::geom_hline(yintercept = mean(wald_df_all$wald_stat_t^2 < unique(wald_df_all$wald_thresh)[1]), color = "blue", linetype = "dashed")
p <- p + ggplot2::geom_vline(xintercept = unique(wald_df_all$wald_thresh)[1], color = "blue", linetype = "dashed")
p <- p + ggplot2::facet_wrap(~contrast_diseases)
p <- p + ggplot2::labs(x = "Wald F-statistic", y = "ECDF", color = "Cell Type")
p <- p + ggplot2::scale_x_continuous(trans="log10")
p <- p + ggplot2::coord_cartesian(xlim = c(1e-3, 1e5))
p <- p + ggplot2::theme_bw()
p <- p + ggplot2::theme(text = element_text(size=20))
p
```

```{r, eval = FALSE}
# PLOT: Wald statistics, as a boxplot

p <- ggplot2::ggplot(wald_df_all, ggplot2::aes(y = wald_stat_t^2,
                                               x = contrast_celltype, 
                                               fill = contrast_diseases))
p <- p + ggplot2::geom_boxplot(outlier.shape = NA)
p <- p + ggplot2::labs(y = "Wald F-statistic", x = "Cell type", fill = "Comparison")
p <- p + ggplot2::scale_y_continuous(trans="log10")
p <- p + ggplot2::coord_cartesian(ylim = c(1e-3, 1e5))
p <- p + ggplot2::theme_bw()
p <- p + ggplot2::theme(text = element_text(size=20), 
                        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
p
```

```{r, eval = FALSE}
# QQ plot of computed Wald p-values (for below-threshold only), faceted by cell type
p <- ggplot2::ggplot(wald_df_all, # %>% dplyr::filter(wald_stat_t < wald_thresh), 
                     ggplot2::aes(sample = wald_pval, color = contrast_diseases))
p <- p + ggplot2::stat_qq(linewidth = 1, distribution = stats::qunif)
p <- p + ggplot2::stat_qq(linewidth = 1, distribution = stats::qunif, color = "black",
                          ggplot2::aes(sample = wald_pval))
p <- p + ggplot2::geom_abline(intercept = 0, slope = 1)
p <- p + ggplot2::facet_wrap(~contrast_celltype)
p <- p + ggplot2::labs(y = "Wald P-Value", x = "U[0, 1]", color = "Comparison")
p <- p + ggplot2::theme_bw()
p <- p + ggplot2::theme(text = element_text(size=20))
p
```

```{r, eval = FALSE}
# QQ plot of computed Wald p-values (for below-threshold only), faceted by disease comparison
p <- ggplot2::ggplot(wald_df_all, # %>% dplyr::filter(wald_stat_t < wald_thresh), 
                     ggplot2::aes(sample = wald_pval, color = contrast_celltype))
p <- p + ggplot2::stat_qq(size = 1 distribution = stats::qunif)
p <- p + ggplot2::stat_qq(size = 1, distribution = stats::qunif, color = "black",
                          ggplot2::aes(sample = wald_pval))
p <- p + ggplot2::geom_abline(intercept = 0, slope = 1)
p <- p + ggplot2::facet_wrap(~contrast_diseases)
p <- p + ggplot2::labs(y = "Wald P-Value", x = "U[0, 1]", color = "Cell Type")
p <- p + ggplot2::theme_bw()
p <- p + ggplot2::theme(text = element_text(size=20))
p
```


```{r, eval = FALSE}
# Dataframe with summary of Wald statistics by comparison
data.frame(wald_df_all 
           %>% dplyr::group_by(fit_model, contrast_celltype, contrast_diseases)
           %>% dplyr::summarise(mean_w=mean(wald_stat_t^2),
                                max_w=max(wald_stat_t^2), 
                                q1_w=quantile(wald_stat_t^2, 0.25), 
                                q2_w=quantile(wald_stat_t^2, 0.5), 
                                q3_w=quantile(wald_stat_t^2, 0.75), 
                                q95_w=quantile(wald_stat_t^2, 0.95)))
```

# Number of significant genes per contrast

```{r}
# Count number of significant genes per contrast
PVAL_THRESH <- 0.05
id_cols <- c("fit_model", 
             "kernel_type", 
             "contrast_indices", 
             "contrast_description"
             )
wald_gene_nsig_df <- data.frame(
  wald_df_all
  %>% dplyr::group_by(across(all_of(id_cols)))
  %>% dplyr::summarise(n_sig_genes = sum(wald_pval_adj < PVAL_THRESH, na.rm =
                                           TRUE))
)

# Create a contrast v. gene matrix of p-values
fit_model_val <- "TESSERA-Leroux"
contrast_gene_mat <- reshape2::acast(
  wald_df_all[(wald_df_all$fit_model == fit_model_val), ],
  contrast_description ~ gene,
  value.var = "wald_pval_adj",
  fun.aggregate = mean
)
contrast_gene_mat_sig <- 1 * (contrast_gene_mat < PVAL_THRESH)
```

We visualize the number of significant genes found via the TESSERA Poisson-Leroux model for each contrast comparison using an FDR of 0.05 with BH-adjusted p-values. 
We see that comparisons involving DKD find markedly more significant genes.

```{r, fig.height=7, fig.width=10}
plot_data <- wald_gene_nsig_df[grepl("-", wald_gene_nsig_df$contrast_description), ]

plot_data$celltype <- unlist(lapply(strsplit(plot_data$contrast_description, ":"),
                                    function (x) x[1]))
plot_data$comparison <- gsub("-", " v. ", unlist(lapply(strsplit(plot_data$contrast_description, ":"),
                                    function (x) x[2])))

p <- ggplot2::ggplot(plot_data, ggplot2::aes(x = celltype, 
                                             y = n_sig_genes,
                                             fill = comparison
                                             ))
p <- p + ggplot2::geom_col(position = position_dodge(preserve = "single", width = 0.9), width = 0.8)
p <- p + ggplot2::scale_fill_manual(values = c("cornflowerblue", "goldenrod1", "hotpink"))
p <- p + ggplot2::labs(x = "Cell Type",
                       y = "Number of significant genes (FDR < 0.05)",
                       fill = "Fitted Model"
                       )
p <- p + ggplot2::theme_bw()
p <- p + ggplot2::theme(text = element_text(size=20))
p <- p + ggplot2::theme(axis.text.x = element_text(size = 10, angle = 90, hjust = 1, vjust=0.5))
p
ggsave(paste0(figure_path, "sig_genes_TESSERA.pdf"), p, width = 8, height=6, units = "in", dpi = 300)
```

```{r}
# Count number of significant genes per contrast
PVAL_THRESH <- 0.05

limma_gene_nsig_df <- data.frame(
  limma_df_all
  %>% dplyr::group_by(across(all_of(c("fit_model", "contrast_description"))))
  %>% dplyr::summarise(n_sig_genes = sum(pval_adj < PVAL_THRESH, na.rm = TRUE))
)

deseq2_gene_nsig_df <- data.frame(
  deseq2_df_all
  %>% dplyr::group_by(across(all_of(c("fit_model", "contrast_description"))))
  %>% dplyr::summarise(n_sig_genes = sum(pval_adj < PVAL_THRESH, na.rm = TRUE))
)

pb_gene_nsig_df <- rbind(deseq2_gene_nsig_df, limma_gene_nsig_df)
```

```{r, fig.height=7, fig.width=15}
pb_plot_data <- pb_gene_nsig_df[grepl("-", pb_gene_nsig_df$contrast_description), ]

pb_plot_data$celltype <- unlist(lapply(strsplit(pb_plot_data$contrast_description, ":"),
                                    function (x) x[1]))
pb_plot_data$comparison <- gsub("-", " v. ", unlist(lapply(strsplit(pb_plot_data$contrast_description, ":"),
                                    function (x) x[2])))

p <- ggplot2::ggplot(pb_plot_data, ggplot2::aes(x = factor(celltype), 
                                             y = n_sig_genes,
                                             fill = comparison
                                             ))
p <- p + ggplot2::facet_wrap(~fit_model)
p <- p + ggplot2::geom_col(position = position_dodge(preserve = "single", width = 0.9), width = 0.8)
p <- p + ggplot2::scale_fill_manual(values = c("cornflowerblue", "goldenrod1", "hotpink"))
p <- p + ggplot2::labs(x = "Cell Type",
                       y = "Number of significant genes (FDR < 0.05)",
                       fill = "Comparison"
                       )
p <- p + ggplot2::theme_bw()
p <- p + ggplot2::theme(text = element_text(size=20))
p <- p + ggplot2::theme(axis.text.x = element_text(size = 10, angle = 90, hjust = 1, vjust=0.5))
p
ggsave(paste0(figure_path, "sig_genes_pseudobulk.pdf"), p, width = 12, height=9, units = "in", dpi = 300)
```

```{r}
glm_wald_df_all$wald_pval <- pchisq(glm_wald_df_all$wald_stat_t^2, 1, lower.tail = FALSE)
glm_wald_df_all$wald_pval_adj <- NA
for (mv in unique(glm_wald_df_all$fit_model)) {
  #for (cn in unique(glm_wald_df_all$contrast_name)) {
    subset_idx <- which((glm_wald_df_all$fit_model == mv) 
                        #& (glm_wald_df_all$contrast_name == cn)
                        )
    glm_wald_df_all$wald_pval_adj[subset_idx] <- p.adjust(glm_wald_df_all$wald_pval[subset_idx], method="BH")
  #}
}
```

```{r}
# Count number of significant genes per contrast
PVAL_THRESH <- 0.05
glm_id_cols <- c("fit_model", "contrast_name")
glm_gene_nsig_df <- data.frame(
  glm_wald_df_all
  %>% dplyr::group_by(across(all_of(glm_id_cols)))
  %>% dplyr::summarise(n_sig_genes = sum(wald_pval_adj < PVAL_THRESH, na.rm =
                                           TRUE))
)
```

We visualize the number of significant genes found via each GLM model (Poisson and Negative Binomial GLMs) for each contrast comparison using an FDR of 0.05 with BH-adjusted p-values. 
We see that comparisons involving DKD find markedly more significant genes, as was also seen with the TESSERA Poisson-Leroux model.

```{r, fig.height=7, fig.width=15}
glm_plot_data <- glm_gene_nsig_df[grepl("-", glm_gene_nsig_df$contrast_name), ]

glm_plot_data$celltype <- unlist(lapply(strsplit(glm_plot_data$contrast_name, ":"),
                                    function (x) x[1]))
glm_plot_data$comparison <- gsub("-", " v. ", unlist(lapply(strsplit(glm_plot_data$contrast_name, ":"),
                                    function (x) x[2])))

p <- ggplot2::ggplot(glm_plot_data, ggplot2::aes(x = celltype, 
                                             y = n_sig_genes,
                                             fill = comparison
                                             ))
p <- p + ggplot2::facet_wrap(~fit_model, ncol=2)
p <- p + ggplot2::geom_col(position = position_dodge(preserve = "single", width = 0.9), width = 0.8)
p <- p + ggplot2::scale_fill_manual(values = c("cornflowerblue", "goldenrod1", "hotpink"))
p <- p + ggplot2::labs(x = "Cell Type",
                       y = "Number of significant genes (FDR < 0.05)",
                       fill=""
                       )
p <- p + ggplot2::theme_bw()
p <- p + ggplot2::theme(text = element_text(size=20), legend.position="bottom")
p <- p + ggplot2::theme(axis.text.x = element_text(size = 10, angle = 90, hjust = 1, vjust=0.5))
p
ggsave(paste0(figure_path, "sig_genes_glm.pdf"), p, width = 12, height=12, units = "in", dpi = 300)
```

```{r}
plot_data_sub <- plot_data[, c("fit_model", "celltype", "n_sig_genes", "comparison")]
glm_plot_data_sub <- glm_plot_data[, c("fit_model", "celltype", "n_sig_genes", "comparison")]
pb_plot_data_sub <- pb_plot_data[, c("fit_model", "celltype", "n_sig_genes", "comparison")]

plot_data_comb <- rbind(plot_data_sub, glm_plot_data_sub, pb_plot_data_sub)
plot_data_comb$fit_model <- factor(plot_data_comb$fit_model, levels=c("TESSERA-Leroux", "DESeq2", "limma-voom", "Poisson GLM", "NB GLM", "GAM"))

model_fill_vals <- c("TESSERA-Leroux" = "#C77CFF", "TESSERA-CAR" = "pink", "TESSERA-SAR" = "indianred1", "TESSERA-spNNGP" = "#32CD32", "DESeq2" = "darkorange", "limma-voom" = "goldenrod1", "Poisson GLM" = "lightblue", "NB GLM" =  "cornflowerblue", "GAM" = "tan4")
```

We see that the TESSERA Poisson-Leroux model finds noticeably fewer significant genes that either GLM model, regardless of the comparison.
This is likely because this model accounts for the spatial correlation inherent to the data and thus is a better fit to the data, thereby potentially leaving us with fewer false positive genes. 
We also see that the Negative Binomial GLM finds fewer significant genes than the Poisson GLM. While the NB GLM does not account for the spatial correlation in the data, it captures some of the overdispersion and is hence less mis-specified than the Poisson GLM.

```{r, fig.height=20, fig.width=18}
p <- ggplot2::ggplot(plot_data_comb, ggplot2::aes(x = celltype, 
                                             y = n_sig_genes,
                                             fill = fit_model
                                             ))
p <- p + ggplot2::facet_wrap(~comparison, ncol = 2)
p <- p + ggplot2::geom_col(position = "dodge")
p <- p + ggplot2::scale_fill_manual(values = model_fill_vals)
p <- p + ggplot2::guides(fill = ggplot2::guide_legend(byrow = TRUE))
p <- p + ggplot2::labs(x = "Cell Type", 
                       y = "Number of significant genes (FDR < 0.05)",
                       fill = "")
p <- p + ggplot2::theme_bw()
p <- p + ggplot2::theme(text = element_text(size=20), legend.position="bottom", legend.key.size = unit(1.3, "cm"), legend.text = element_text(size = 22))
p <- p + ggplot2::theme(axis.title.x = ggplot2::element_text(size = 25), axis.title.y = ggplot2::element_text(size = 25))
p <- p + ggplot2::theme(axis.text.x = element_text(size = 20, angle = 90, hjust = 1, vjust=0.5))
p <- p + ggplot2::theme(strip.text = ggplot2::element_text(size = 25))
p
ggsave(paste0(figure_path, "sig_genes_all_methods_by_contrast.pdf"), p, width = 16, height=16, units = "in", dpi = 300)
```


```{r, fig.height=20, fig.width=18}
plot_data_comb_sub <- plot_data_comb[which(plot_data_comb$comparison=="Control v. DKD" & plot_data_comb$celltype %in% c("C_TAL", "PT_S1", "Fib", "Podo")), ]
p <- ggplot2::ggplot(plot_data_comb_sub, ggplot2::aes(x = celltype, 
                                             y = n_sig_genes,
                                             fill = fit_model
                                             ))
p <- p + ggplot2::geom_col(position = "dodge")
p <- p + ggplot2::scale_fill_manual(values = model_fill_vals)
p <- p + ggplot2::labs(x = "Cell Type", 
                       y = "Number of significant genes (FDR < 0.05)",
                       fill = "Fitted Model")
p <- p + ggplot2::theme_bw()
p <- p + ggplot2::theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust=0.5))
p <- p + ggplot2::theme(text = element_text(size=20))
p
ggsave(paste0(figure_path, "sig_genes_all_methods_Control_v_DKD.pdf"), p, width = 9, height=8, units = "in", dpi = 300)
```

We notice once more that the TESSERA Poisson-Leroux model finds noticeably fewer significant genes that either GLM model. We also see that, regardless of the model type, all comparisons involving DKD find markedly more significant genes.

```{r, fig.height=20, fig.width=18}
p <- ggplot2::ggplot(plot_data_comb, ggplot2::aes(x = celltype, 
                                             y = n_sig_genes,
                                             fill = comparison
                                             ))
p <- p + ggplot2::facet_wrap(~fit_model, ncol = 2)
p <- p + ggplot2::geom_col(position = position_dodge(preserve = "single", width = 0.9), width = 0.8)
p <- p + ggplot2::scale_fill_manual(values = c("cornflowerblue", "goldenrod1", "hotpink", "seagreen", "purple"))
p <- p + ggplot2::labs(x = "Cell Type", 
                       y = "Number of significant genes (FDR < 0.05)",
                       fill = "")
p <- p + ggplot2::theme_bw()
p <- p + ggplot2::theme(text = element_text(size=20), legend.position="bottom", legend.key.size = unit(1.3, "cm"), legend.text = element_text(size = 22))
p <- p + ggplot2::theme(axis.title.x = ggplot2::element_text(size = 25), axis.title.y = ggplot2::element_text(size = 25))
p <- p + ggplot2::theme(axis.text.x = element_text(size = 20, angle = 90, hjust = 1, vjust=0.5))
p <- p + ggplot2::theme(strip.text = ggplot2::element_text(size = 25))
p
ggsave(paste0(figure_path, "sig_genes_all_contrasts_by_method.pdf"), p, width = 15, height=15, units = "in", dpi = 300)
```

We see for the TESSERA Poisson-Leroux model all contrast comparisons ordered by number of associated significant genes. We can again see that comparisons involving DKD find markedly more genes than the comparison not involving DKD. We also see that cell types with low numbers of cells, such as RBC, find fewer significant genes, as expected.

```{r, fig.height=7, fig.width=18}
p <- ggplot2::ggplot(plot_data, ggplot2::aes(x = fct_reorder(contrast_description,
                                                             -n_sig_genes), 
                                             y = n_sig_genes,
                                             fill = comparison
                                             ))
p <- p + ggplot2::geom_col(position = "dodge")
p <- p + ggplot2::scale_fill_manual(values = c("cornflowerblue", "goldenrod1", "hotpink"))
p <- p + ggplot2::labs(x = "Cell Type", 
                       y = "Number of sig. genes (FDR < 0.05)",
                       fill = "Comparison"
                       )
p <- p + ggplot2::theme_bw()
p <- p + ggplot2::theme(text = element_text(size=20))
p <- p + ggplot2::theme(axis.text.x = element_text(size=8, angle = 90, hjust = 1, vjust=0.5))
p
ggsave(paste0(figure_path, "sig_genes_TESSERA_ordered.pdf"), p, width = 12, height=9, units = "in", dpi = 300)
```

# Jaccard similarity

```{r}
# Ranks
wald_df_all$gene_rank <- NA
for (mv in unique(wald_df_all$fit_model)) {
  for (cd in unique(wald_df_all$contrast_description)) {
    tmp_idx <- which((wald_df_all$fit_model == mv) & (wald_df_all$contrast_description == cd))
    wald_df_all$gene_rank[tmp_idx] <- length(tmp_idx) - rank(wald_df_all$wald_stat_t[tmp_idx]^2) + 1
  }
}
wald_df_all$contrast_name <- wald_df_all$contrast_description

glm_wald_df_all$gene_rank <- NA
for (mv in unique(glm_wald_df_all$fit_model)) {
  for (cd in unique(glm_wald_df_all$contrast_name)) {
    tmp_idx <- which((glm_wald_df_all$fit_model == mv) & (glm_wald_df_all$contrast_name == cd))
    glm_wald_df_all$gene_rank[tmp_idx] <- length(tmp_idx) - rank(glm_wald_df_all$wald_stat_t[tmp_idx]^2) + 1
  }
}
glm_wald_df_all$contrast_description <- glm_wald_df_all$contrast_name

limma_df_all$gene_rank <- NA
for (mv in unique(limma_df_all$fit_model)) {
  for (cd in unique(limma_df_all$contrast_description)) {
    tmp_idx <- which((limma_df_all$fit_model == mv) & (limma_df_all$contrast_description == cd))
    limma_df_all$gene_rank[tmp_idx] <- length(tmp_idx) - rank(limma_df_all$t[tmp_idx]) + 1
  }
}
limma_df_all$contrast_name <- limma_df_all$contrast_description

deseq2_df_all$gene_rank <- NA
for (mv in unique(deseq2_df_all$fit_model)) {
  for (cd in unique(deseq2_df_all$contrast_description)) {
    tmp_idx <- which((deseq2_df_all$fit_model == mv) & (deseq2_df_all$contrast_description == cd))
    deseq2_df_all$gene_rank[tmp_idx] <- length(tmp_idx) - rank(deseq2_df_all$stat[tmp_idx]) + 1
  }
}
deseq2_df_all$contrast_name <- deseq2_df_all$contrast_description
```


```{r}
# Jaccard similarity of top-n genes for each pair of methods and contrast
tmp_wald_df_all <- dplyr::bind_rows(glm_wald_df_all, wald_df_all, limma_df_all, deseq2_df_all)
n_list <- seq(1, 3000)

# Function to compute Jaccard for two methods at different n values
compute_jaccard <- function(df, mv1, mv2, n_list) {
  map_dfr(n_list, function(n) {
    g1 <- df %>% filter(fit_model == mv1, gene_rank <= n) %>% pull(gene)
    g2 <- df %>% filter(fit_model == mv2, gene_rank <= n) %>% pull(gene)
    n_inter <- length(intersect(g1, g2))
    n_union <- length(unique(c(g1, g2)))
    tibble(
      fit_model1 = mv1,
      fit_model2 = mv2,
      n = n,
      n_intersect = n_inter,
      n_union = n_union,
      jaccard = n_inter / n_union
    )
  })
}

# Do this for each contrast
plot_df <- tmp_wald_df_all %>%
  dplyr::filter(contrast_name == "PT_S1:Control-DKD") %>%
  split(.$contrast_name) %>%
  imap_dfr(function(df, contrast_name) {
    # All unique model pairs
    model_pairs <- combn(unique(df$fit_model), 2, simplify = FALSE)
    map_dfr(model_pairs, ~compute_jaccard(df, .x[1], .x[2], n_list)) %>%
      mutate(contrast_description = contrast_name)
  })
```


```{r, eval = FALSE}
# Jaccard similarity of top-n genes for each pair of methods and contrast
# tmp_wald_df_all <- dplyr::bind_rows(glm_wald_df_all, wald_df_all, limma_df_all, deseq2_df_all)
# tmp_wald_df <- tmp_wald_df_all[which(tmp_wald_df_all$contrast_description == "PT_S1:Control-DKD"), ]
# n_list <- seq(1, 3000)
# 
# plot_df <- list()
# 
# for (mv1 in unique(tmp_wald_df$fit_model)) { # Loop over method 1
#   for (mv2 in unique(tmp_wald_df$fit_model)) { # Loop over method 2
#     if (mv1 == mv2) { # Don't bother with the same same method (trivial)
#       next
#     } else if (mv2 < mv1) { # We don't need to compare both (x, y) and (y, x)
#       next
#     }
# 
#     for (cd in unique(tmp_wald_df$contrast_description)) { # Loop over contrasts
#       #cat(Sys.time(), mv1, mv2, cd, "\n")
#       for (idx in 1:length(n_list)) {
#         g1 <- tmp_wald_df$gene[(tmp_wald_df$gene_rank <= n_list[idx]) & (tmp_wald_df$fit_model == mv1)]
#         g2 <- tmp_wald_df$gene[(tmp_wald_df$gene_rank <= n_list[idx]) & (tmp_wald_df$fit_model == mv2)]
# 
#         plot_df[[1 + length(plot_df)]] <- data.frame(
#           fit_model1 = mv1,
#           fit_model2 = mv2,
#           contrast_description = cd,
#           n = n_list[idx],
#           n_intersect = length(intersect(g1, g2)),
#           n_union = length(unique(c(g1, g2)))
#         )
#       }
#     }
#   }
# }
# plot_df <- dplyr::bind_rows(plot_df)
# plot_df$jaccard <- plot_df$n_intersect / plot_df$n_union
```

```{r}
# Expected Jaccard similarity of random ordering
j_trials <- 100
j_mat <- array(data=NA, dim=c(length(n_list), j_trials, 2))
gene_list <- unique(tmp_wald_df_all$gene)
for (tr in 1:j_trials) {
  g1 <- sample(gene_list, replace = FALSE)
  g2 <- sample(gene_list, replace = FALSE)

  for (idx in 1:nrow(j_mat)) {
    j_mat[idx, tr, 1] <- length(intersect(g1[1:n_list[idx]], g2[1:n_list[idx]]))
    j_mat[idx, tr, 2] <- length(union(g1[1:n_list[idx]], g2[1:n_list[idx]]))
  }
}
exp_jaccard_data <- data.frame(n=n_list, expected_jaccard=rowMeans(j_mat[, , 1] / j_mat[, , 2]))

plot_df <- merge(plot_df, exp_jaccard_data, by="n")
```


```{r, fig.height=8, fig.width=10}
plot_df$method_comparison <- paste0(plot_df$fit_model1, " v. ", plot_df$fit_model2)
# keep_comp <- c("GLM_NB v. GLM_Poisson", "DESeq2 v. limma_voom",
#                "GLM_Poisson v. TESSERA_Leroux", "GLM_NB v. TESSERA_Leroux",
#                "limma_voom v. TESSERA_Leroux", "DESeq2 v. TESSERA_Leroux")
# plot_df <- plot_df[which(plot_df$method_comparison %in% keep_comp), ]
# plot_df$method_comparison <- factor(plot_df$method_comparison, levels = keep_comp)
plot_df$method_comparison <- factor(plot_df$method_comparison, 
                                    levels = c("limma-voom v. DESeq2",
                                               "NB GLM v. Poisson GLM",
                                               "GAM v. NB GLM", 
                                               "GAM v. Poisson GLM",
                                               "NB GLM v. DESeq2", 
                                               "NB GLM v. limma-voom",
                                               "Poisson GLM v. DESeq2",
                                               "Poisson GLM v. limma-voom",
                                               "GAM v. DESeq2",
                                               "GAM v. limma-voom",
                                               "TESSERA-Leroux v. DESeq2",
                                               "TESSERA-Leroux v. limma-voom",
                                               "NB GLM v. TESSERA-Leroux",
                                               "Poisson GLM v. TESSERA-Leroux", 
                                               "GAM v. TESSERA-Leroux"
                                               )
                                    )
p1 <- ggplot2::ggplot(data=plot_df, ggplot2::aes(x=n, 
                                                 y=jaccard,
                                                 color=method_comparison
                                                 ))
p1 <- p1 + ggplot2::geom_point(size=1)
p1 <- p1 + ggplot2::geom_line(ggplot2::aes(x=n, y=expected_jaccard), color="black", lwd=1.5)
p1 <- p1 + facet_wrap(~contrast_description)
p1 <- p1 + scale_color_manual(values = c("sienna", "darkgray", "pink", "red", "gold", "orange", "lightgreen", "forestgreen","cornflowerblue", "blue", "violet", "purple", "turquoise", "dodgerblue", "tan"))
p1 <- p1 + ggplot2::guides(color = guide_legend(override.aes = list(size = 3)))
p1 <- p1 + ggplot2::labs(x = expression("Top " * italic(n) * " Genes"), 
                         y = "Jaccard Similarity",
                         color = "Comparison"
                         )
p1 <- p1 + ggplot2::theme_bw()
p1 <- p1 + ggplot2::theme(text=ggplot2::element_text(size=20),
                          axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
p1
ggsave(paste0(figure_path, "topngenes_PT_S1_Control_DKD.pdf"), p1, width = 12, height=9, units = "in")
```

