---
title: "Synthetic data model performance"
author: "Florica"
date: "2025-07-04"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r}
# Library
library(dplyr)
library(forcats)
library(ggplot2)
library(patchwork)
library(rstatix)

# Paths
processed_data_path <- "../../data/processed/"
figure_path <- "../../figures/synth/"

# Model type
pattern_type <- "spNNGP" # "spNNGP" "Leroux"

# Read in kidney data
seu_kid <- readRDS(paste0(processed_data_path, "seu_kidney.rds"))
seu_kid@meta.data$sample <- seu_kid@meta.data$orig.ident

# Genes ordered by percentage of cells expressed
uniq_samp <- unique(seu_kid@meta.data$sample)
pct_xpr_df <- list()
for(s_idx in 1:length(uniq_samp)){
  seu_kid_sub <- seu_kid[which(seu_kid@meta.data$sample == uniq_samp[s_idx]), ]
  pct_xpr_df[[s_idx]] <- data.frame(gene = rownames(seu_kid_sub@assays$RNA@features),
                           sample = uniq_samp[s_idx],
                           pct_xpr = Matrix::rowSums(seu_kid_sub@assays$RNA@layers$counts > 0) / ncol(seu_kid_sub)
) 
}
pct_xpr_df <- dplyr::bind_rows(pct_xpr_df)

# Non-spatial results
nonspat_summary_out_all <- readRDS(paste0(processed_data_path, "nonspatial_results_perfsummary_onlyinteract.rds"))
nonspat_glm_gam_summary_out_all <- readRDS(paste0(processed_data_path, "nonspatial_gam_glm_perfsummary_onlyinteract.rds"))
nonspat_glm_gam_summary_out_all$Moran_residuals <- nonspat_glm_gam_summary_out_all$Moran_resid

nonspat_summary_out_all <- dplyr::bind_rows(nonspat_summary_out_all, nonspat_glm_gam_summary_out_all)
nonspat_summary_out_all <- nonspat_summary_out_all %>%
  mutate(fit_model = recode(fit_model,
                            "Leroux" = "TESSERA-Leroux",
                            "CAR" = "TESSERA-CAR",
                            "SAR" = "TESSERA-SAR",
                            "spNNGP" = "TESSERA-spNNGP",
                            "Poisson_GLM" = "Poisson GLM",
                            "NB_GLM" = "NB GLM",
                            "GAM" = "GAM"))
nonspat_summary_out_all$fit_model <- factor(nonspat_summary_out_all$fit_model, levels = c("TESSERA-Leroux", "TESSERA-CAR", "TESSERA-SAR", "TESSERA-spNNGP", "Poisson GLM", "NB GLM", "GAM"))
nonspat_summary_out_all$gene <- factor(nonspat_summary_out_all$gene, levels = c("IGKC", "DDX24", "SMTN", "CRLS1", "ERP44"))

summary_out_all <- readRDS(paste0(processed_data_path, "synth_results_perfsummary_onlyinteract.rds"))
summary_out_all <- dplyr::left_join(summary_out_all, dplyr::distinct(seu_kid@meta.data[, c("sample", "Group")]), by = "sample")

glm_summary_out_all <- readRDS(paste0(processed_data_path, "glm_gam_synth_results_perfsummary_onlyinteract.rds"))
                                   
# Add in diagnosis information
summary_out_all$Group <- factor(summary_out_all$Group, levels = c("Control", "HKD", "DKD"))
summary_out_all$sample <- forcats::fct_reorder(summary_out_all$sample, as.numeric(summary_out_all$Group))

summary_out_all <- summary_out_all[which(summary_out_all$gene %in% c("IGKC", "DDX24", "SMTN", "CRLS1", "ERP44")), ]
summary_out_all$gene <- factor(summary_out_all$gene, levels = c("IGKC", "DDX24", "SMTN", "CRLS1", "ERP44"))

# Rename models
summary_out_all <- dplyr::bind_rows(summary_out_all, glm_summary_out_all)
summary_out_all <- summary_out_all %>%
  mutate(fit_model = recode(fit_model,
                            "Leroux" = "TESSERA-Leroux",
                            "CAR" = "TESSERA-CAR",
                            "SAR" = "TESSERA-SAR",
                            "spNNGP" = "TESSERA-spNNGP",
                            "Poisson_GLM" = "Poisson GLM",
                            "NB_GLM" = "NB GLM",
                            "GAM" = "GAM"))
summary_out_all$fit_model <- factor(summary_out_all$fit_model, levels = c("TESSERA-Leroux", "TESSERA-CAR", "TESSERA-SAR", "TESSERA-spNNGP", "Poisson GLM", "NB GLM", "GAM"))
summary_out_all$gene <- factor(summary_out_all$gene, levels = c("IGKC", "DDX24", "SMTN", "CRLS1", "ERP44"))

summary_out <- summary_out_all[which(summary_out_all$data_model == pattern_type), ]
nonspat_summary_out <- nonspat_summary_out_all[which(nonspat_summary_out_all$datagen_model == pattern_type), ]
```

```{r}
# Colors for models when using color and fill in ggplots
model_color_vals <- c("TESSERA-Leroux" = "#800080", "TESSERA-CAR" = "hotpink4", "TESSERA-SAR" = "darkred", "TESSERA-spNNGP" = "#006400", "Poisson GLM" = "steelblue4", "NB GLM" =  "#00008B", "GAM" = "tan4")
model_fill_vals <- c("TESSERA-Leroux" = "#C77CFF", "TESSERA-CAR" = "pink", "TESSERA-SAR" = "indianred1", "TESSERA-spNNGP" = "#32CD32", "Poisson GLM" = "lightblue", "NB GLM" =  "cornflowerblue", "GAM" = "burlywood")

sample_color_vals <- c("#d62728","#1f77b4","#2ca02c","#ff7f0e","#9467bd","#8c564b","#e377c2",
                       "#7f7f7f","#bcbd22","#17becf","#ff9896","#aec7e8","#98df8a","#ffbb78")
```

# Relative squared error in beta

```{r}
plot_df <- summary_out_all
plot_df <- plot_df[which(plot_df$gene %in% c("IGKC", "ERP44")), ]
plot_df <- plot_df %>%
  mutate(alt_gene_name = recode(gene,
                            "IGKC" = "High",
                            "ERP44" = "Low"))
p <- ggplot2::ggplot(plot_df, aes(x = alt_gene_name, y = SE_beta / Norm2_beta, fill = fit_model, color = fit_model))
p <- p + ggplot2::geom_boxplot(outliers = FALSE)
p <- p + ggplot2::facet_wrap(~paste0("Poisson-", data_model))
p <- p + ggplot2::scale_color_manual(name = "", values = model_color_vals)
p <- p + ggplot2::scale_fill_manual(name = "", values = model_fill_vals)
p <- p + ggplot2::guides(fill = ggplot2::guide_legend(byrow = TRUE))
p <- p + ggplot2::guides(color = ggplot2::guide_legend(byrow = TRUE))
p <- p + ggplot2::scale_y_log10()
p <- p + ggplot2::labs(x = "Gene Variance", y = expression(paste("Relative squared error in ", beta)))
p <- p + ggplot2::theme_bw()
p <- p + ggplot2::theme(text = element_text(size=20), legend.position = "bottom")
p
ggplot2::ggsave(paste0(figure_path, "synth_beta_error.pdf"), p, width=10, height=7, units = "in", dpi = 300)
```

# Relative squared error in beta

```{r}
p <- ggplot2::ggplot(summary_out_all, aes(x = gene, y = SE_beta / Norm2_beta, fill = fit_model, color = fit_model))
p <- p + ggplot2::geom_boxplot(outliers = FALSE)
p <- p + ggplot2::facet_wrap(~paste0("Poisson-", data_model))
p <- p + ggplot2::scale_color_manual(name = "", values = model_color_vals)
p <- p + ggplot2::scale_fill_manual(name = "", values = model_fill_vals)
p <- p + ggplot2::guides(fill = ggplot2::guide_legend(byrow = TRUE))
p <- p + ggplot2::guides(color = ggplot2::guide_legend(byrow = TRUE))
p <- p + ggplot2::scale_y_log10()
p <- p + ggplot2::scale_x_discrete(labels = c("Maximum", expression(Q[3]), "Median", expression(Q[1]), "Minimum"))
p <- p + ggplot2::labs(x = "Gene Variance", y = expression(paste("Relative squared error in ", beta)))
p <- p + ggplot2::theme_bw()
p <- p + ggplot2::theme(text = element_text(size=20), legend.position = "bottom")
p
ggplot2::ggsave(paste0(figure_path, "synth_beta_error_all.pdf"), p, width=10, height=7, units = "in", dpi = 300)
```

\newpage

# Performance metrics: Fit quality 

Normalized MSE: MSE / mean([counts]^2), where we define: 
MSE: mean([counts - estimated counts]^2)

Plot shows the normalized MSE where each point corresponds to one gene-sample pair.

```{r}
plot_data <- summary_out_all[which(summary_out_all$data_model=="Leroux"), ]
p1 <- ggplot2::ggplot(plot_data, aes(x = gene, y = MSE_counts_total / Mean_counts2_total, fill = fit_model, color = fit_model))
p1 <- p1 + ggplot2::geom_boxplot(outliers = FALSE)
p1 <- p1 + ggplot2::scale_color_manual(name = "", values = model_color_vals)
p1 <- p1 + ggplot2::scale_fill_manual(name = "", values = model_fill_vals)
p1 <- p1 + ggplot2::guides(fill = ggplot2::guide_legend(byrow = TRUE))
p1 <- p1 + ggplot2::guides(color = ggplot2::guide_legend(byrow = TRUE))
p1 <- p1 + ggplot2::facet_wrap(~paste0("Poisson-", data_model))
p1 <- p1 + ggplot2::scale_y_log10(breaks=c(1e-5, 1e-4, 1e-3, 2e-1, 0.3, 1.0, 3.0))
p1 <- p1 + ggbreak::scale_y_break(breaks=c(1e-3, 2e-1), scales = 5)
p1 <- p1 + ggplot2::labs(x = "Gene", y = "Normalized MSE of counts")
p1 <- p1 + ggplot2::theme_bw()
p1 <- p1 + ggplot2::theme(text = element_text(size=20),
                        legend.position = "bottom",
                        axis.text.y.right = element_blank(),
                        axis.line.y.right = element_blank(),
                        axis.ticks.y.right = element_blank())
p1 <- p1 + theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
ggplot2::ggsave(paste0(figure_path, "synth_counts_error_Leroux.pdf"), p1, width=10, height=7, units = "in", dpi = 300, onefile=FALSE)
```

```{r}
plot_data <- summary_out_all[which(summary_out_all$data_model=="spNNGP"), ]
p2 <- ggplot2::ggplot(plot_data, aes(x = gene, y = MSE_counts_total / Mean_counts2_total, fill = fit_model, color = fit_model))
p2 <- p2 + ggplot2::geom_boxplot(outliers = FALSE)
p2 <- p2 + ggplot2::scale_color_manual(name = "", values = model_color_vals)
p2 <- p2 + ggplot2::scale_fill_manual(name = "", values = model_fill_vals)
p2 <- p2 + ggplot2::guides(fill = ggplot2::guide_legend(byrow = TRUE))
p2 <- p2 + ggplot2::guides(color = ggplot2::guide_legend(byrow = TRUE))
p2 <- p2 + ggplot2::facet_wrap(~paste0("Poisson-", data_model))
p2 <- p2 + ggplot2::ylim(c(0, 7))
p2 <- p2 + ggplot2::labs(x = "Gene", y = "Normalized MSE of counts")
p2 <- p2 + ggplot2::theme_bw()
p2 <- p2 + ggplot2::theme(text = element_text(size=20),
                        legend.position = "bottom",
                        axis.text.y.right = element_blank(),
                        axis.line.y.right = element_blank(),
                        axis.ticks.y.right = element_blank())
p2 <- p2 + theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
ggplot2::ggsave(paste0(figure_path, "synth_counts_error_spNNGP.pdf"), p2, width=8, height=6, units = "in", dpi = 300, onefile=FALSE)
```

```{r}
# Set shared theme elements to ensure identical spacing
shared_theme <- theme_bw() + 
  theme(text = element_text(size = 20),
        #axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none",
        axis.title.x = element_blank()) # Remove "Gene" from individual plots

# Leroux Column (Stacked manually to avoid ggbreak bugs)
p1_top <- ggplot(subset(summary_out_all, data_model == "Leroux"), 
                 aes(x = gene, y = MSE_counts_total / Mean_counts2_total, fill = fit_model, color = fit_model)) +
  geom_boxplot(outliers = FALSE, width = 0.7) +
  ggplot2::scale_x_discrete(labels = c("Maximum", expression(Q[3]), "Median", expression(Q[1]), "Minimum")) +
  scale_y_log10(limits = c(0.2, 3.5), breaks = c(0.2, 0.3, 1.0, 3.0), labels = c("0.2", "0.3", "1.0", "3.0")) +
  facet_wrap(~"Poisson-Leroux") +
  scale_color_manual(values = model_color_vals) +
  scale_fill_manual(values = model_fill_vals) +
  shared_theme +
  theme(text = element_text(size=20), axis.text.x = element_blank(), axis.ticks.x = element_blank())

p1_bot <- ggplot(subset(summary_out_all, data_model == "Leroux"), 
                 aes(x = gene, y = MSE_counts_total / Mean_counts2_total, fill = fit_model, color = fit_model)) +
  geom_boxplot(outliers = FALSE, width = 0.7) +
  ggplot2::scale_x_discrete(labels = c("Maximum", expression(Q[3]), "Median", expression(Q[1]), "Minimum")) +
  scale_y_log10(limits = c(1e-5, 1.2e-3), breaks = c(1e-5, 1e-4, 1e-3)) +
  scale_color_manual(values = model_color_vals) +
  scale_fill_manual(values = model_fill_vals) +
  shared_theme + 
  theme(strip.background = element_blank(), strip.text = element_blank())

# spNNGP Plot (Ensure it has exactly the same shared_theme)
p2 <- ggplot(subset(summary_out_all, data_model == "spNNGP"), 
             aes(x = gene, y = MSE_counts_total / Mean_counts2_total, fill = fit_model, color = fit_model)) +
  geom_boxplot(outliers = FALSE, width = 0.7) +
  ggplot2::scale_x_discrete(labels = c("Maximum", expression(Q[3]), "Median", expression(Q[1]), "Minimum")) +
  scale_y_continuous(breaks = seq(0, 7, by = 2), limits = c(0, 7)) +
  facet_wrap(~"Poisson-spNNGP") +
  scale_color_manual(values = model_color_vals) +
  scale_fill_manual(values = model_fill_vals) +
  shared_theme

# Ensure subplots are clean (No titles)
p1_top <- p1_top + labs(x = NULL, y = NULL)
p1_bot <- p1_bot + labs(x = NULL, y = NULL)
p2     <- p2     + labs(x = NULL, y = NULL)

# We remove the bottom margin of the top plot and the top margin of the bottom plot
p1_top <- p1_top + theme(plot.margin = margin(b = -20, unit = "pt"))
p1_bot <- p1_bot + theme(plot.margin = margin(t = -20, unit = "pt"))
# Assemble the stack with zero spacing
# Using unit(0, "pt") is more precise than "lines"
p1_stack <- (p1_top / p1_bot) + 
  plot_layout(heights = c(3, 1)) & 
  theme(panel.spacing.y = unit(0, "pt"), 
        plot.margin = margin(0, 0, 0, 0))

# Reduce the left margin (changed from 50 to 35)
p_body <- (p1_stack | p2) + 
  plot_layout(guides = "collect") & 
  theme(legend.position = "bottom", 
        legend.title = element_blank(),
        text = element_text(size = 20),
        # Adjusted left margin to 35 for tighter spacing
        plot.margin = margin(10, 10, 2, 30),
        panel.spacing.y = unit(-1.5, "lines")) 

# Use ggdraw with a slightly higher x-value to bring label closer
p_fin <- cowplot::ggdraw(p_body) + 
  cowplot::draw_label("Normalized MSE of counts", 
                      x = 0.045,     # Increased from 0.03 to move closer to ticks
                      y = 0.60,      # Kept at 0.60 to stay centered on boxes
                      angle = 90, 
                      fontface = "plain", 
                      size = 20) +
  cowplot::draw_label("Gene Variance", 
                      x = 0.53,      
                      y = 0.105, # Move the x-axis label (the smaller, the further down)   
                      fontface = "plain", 
                      size = 20)

# Save
ggsave(paste0(figure_path, "synth_counts_error_all_final.pdf"), 
       p_fin, width = 12, height = 8, units = "in", dpi = 300)
```


\newpage

# Performance metrics: Fit quality 

How the Moran's I of the raw count inputs compares to Moran's I of the residuals. 
We see that almost all gene-sample pairs have a strong decrease for the lattice models. 
Note that the expected value of Moran's I under the null hypothesis of no spatial autocorrelation is 1 / (Number of cells - 1). 

```{r, eval=FALSE}
tmp_df <- summary_out
tmp_df <- (tmp_df 
           %>% dplyr::group_by(gene, sample, fit_model) 
           %>% dplyr::summarize(prop_MI_decr=mean(Moran_residuals < Moran_counts, na.rm=TRUE),
                                   MI_drop=mean(Moran_residuals - Moran_counts, na.rm=TRUE),
                                   MI_resid=mean(Moran_residuals, na.rm=TRUE)
                                )
)
```

```{r, fig.width = 25, fig.height=20}
# Moran counts v. Moran residuals
p <- ggplot2::ggplot(nonspat_summary_out, aes(x = Moran_counts, y = Moran_residuals))
p <- p + ggplot2::geom_point(alpha = 0.5)
p <- p + ggplot2::facet_wrap(~fit_model)
p <- p + ggplot2::labs(x = "Moran's I(counts)", y = "Moran's I(residuals)")
p <- p + ggplot2::theme_bw()
p <- p + ggplot2::theme(text = element_text(size=20))
p
ggplot2::ggsave(paste0(figure_path, "morans_nonspatial_residuals_", pattern_type, ".pdf"), p, width=12, height=9, units = "in", dpi = 300)
```

```{r, fig.width = 25, fig.height=20}
# Moran counts v. Moran residuals
plot_df <- nonspat_summary_out[which(nonspat_summary_out$fit_model %in% c("TESSERA-Leroux", "Poisson GLM", "NB GLM", "GAM")), ]
p <- ggplot2::ggplot(plot_df, aes(x = fit_model, y = Moran_residuals, fill=fit_model, color=fit_model))
p <- p + ggplot2::geom_boxplot(outliers=FALSE)
p <- p + ggplot2::scale_color_manual(name = "", values = model_color_vals)
p <- p + ggplot2::scale_fill_manual(name = "", values = model_fill_vals)
p <- p + ggplot2::labs(x = "Fitted Model", y = "Moran's I(residuals)")
p <- p + ggplot2::theme_bw()
p <- p + ggplot2::theme(text = element_text(size=20), legend.position = "none")
p
ggplot2::ggsave(paste0(figure_path, "morans_nonspatial_residuals_", pattern_type, "_boxplot.pdf"), p, width=10, height=7, units = "in", dpi = 300)
```

```{r, fig.width = 25, fig.height=20}
# Moran counts v. Moran residuals
plot_df <- summary_out[which(summary_out$fit_model %in% c("TESSERA-Leroux", "Poisson GLM", "NB GLM", "GAM")), ]
p <- ggplot2::ggplot(summary_out, aes(x = Moran_counts, y = Moran_residuals - Moran_counts))
p <- p + ggplot2::geom_point(alpha = 0.5)
p <- p + ggplot2::facet_wrap(~fit_model)
p <- p + ggplot2::labs(x = "Moran's I(counts)", y = "Moran's I(residuals) - Moran's I(counts)")
p <- p + ggplot2::theme_bw()
p <- p + ggplot2::theme(text = element_text(size=20))
p
ggplot2::ggsave(paste0(figure_path, "morans_residuals_", pattern_type, ".pdf"), p, width=12, height=9, units = "in", dpi = 300)
```

```{r}
nonspat_summary_out$data_type <- "nonspatial"
nonspat_summary_out$data_model <- nonspat_summary_out$datagen_model 
summary_out$data_type <- "spatial"
summary_out[is.na(summary_out$trial), ]$trial <- summary_out[is.na(summary_out$trial), ]$trial_idx


out_all <- dplyr::bind_rows(nonspat_summary_out, summary_out)
cols_keep_out <- c(colnames(summary_out)[1:6], "data_type", 
                   "Moran_counts", "Moran_residuals",
                   "MSE_counts_sample", "Mean_counts2_sample",
                   "MSE_counts_total", "Mean_counts2_total"
                   )
out_all <- out_all[, cols_keep_out]
n_cells_df <- dplyr::distinct(out_all[, c("sample", "n_cells")])
n_cells_df <- n_cells_df[!is.na(n_cells_df$n_cells), ]
out_all <- merge(out_all, n_cells_df, by="sample", suffixes = c("", "_new"))
out_all$n_cells <- out_all$n_cells_new

# Define the signed square root function
# It takes the sqrt of the absolute value, then restores the sign
signed_sqrt <- function(x) {
  sign(x) * sqrt(abs(x))
}
# Define the inverse function
# It squares the absolute value, then restores the sign
inverse_signed_sqrt <- function(x) {
  sign(x) * (x^2)
}

# Create the new transformation object
# You need to provide a name, the transform, and the inverse transform
signed_sqrt_trans <- scales::trans_new(
  name = "signed_sqrt_trans",
  transform = signed_sqrt,
  inverse = inverse_signed_sqrt
)
```


```{r}
p <- ggplot2::ggplot(out_all, aes(x = fit_model, y = Moran_residuals, fill = data_type))
p <- p + ggplot2::geom_boxplot(outliers=FALSE)
p <- p + ggplot2::geom_hline(yintercept = 0, color = "red", linetype = "dashed", size = 1) # Add line at y=0
p <- p + ggplot2::scale_y_continuous(trans = signed_sqrt_trans)
p <- p + ggplot2::labs(x = "Model", y = "Moran's I", fill = "Simulation Type")
p <- p + ggplot2::theme_bw()
p <- p + ggplot2::theme(text = element_text(size=20), axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
p
```


```{r}
out_all$expected_MI <- -1.0 / (out_all$n_cells - 1)
p <- ggplot2::ggplot(out_all, aes(x = fit_model, y = Moran_residuals - expected_MI, fill = data_type))
p <- p + ggplot2::geom_boxplot(outliers=FALSE)
p <- p + ggplot2::geom_hline(yintercept = 0, color = "red", linetype = "dashed", size = 1) # Add line at y=0
p <- p + ggplot2::scale_y_continuous(trans = signed_sqrt_trans)
p <- p + ggplot2::labs(x = "Model", y = "Moran's I - E[Moran's I]", fill = "Simulation Type")
p <- p + ggplot2::theme_bw()
p <- p + ggplot2::theme(text = element_text(size=20), axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
p
ggplot2::ggsave(paste0(figure_path, "morans_pair_boxplot_", pattern_type, ".pdf"), p, width=12, height=9, units = "in", dpi = 300)
```

```{r}
# Clean the data to ensure 1-to-1 pairing
out_cleaned <- out_all %>%
  # Group by everything that defines a unique observation
  group_by(data_model, fit_model, trial, gene, sample, data_type) %>%
  summarise(Moran_residuals = mean(Moran_residuals, na.rm = TRUE), .groups = "drop") %>%
  
  # Now filter for groups that have exactly both data_types
  group_by(data_model, fit_model, trial, gene, sample) %>%
  filter(n_distinct(data_type) == 2) %>%
  ungroup()

# Run the paired test
paired_results <- out_cleaned %>%
  group_by(data_model, fit_model) %>%
  # Arrange so data_type levels (e.g., A then B) are identical for all pairs
  arrange(data_model, fit_model, trial, gene, sample, data_type) %>%
  wilcox_test(
    Moran_residuals ~ data_type, 
    paired = TRUE
  ) %>%
  add_significance()

print(paired_results)
```

```{r}
unpaired_results <- out_all %>%
  group_by(data_model, fit_model) %>%
  # Filter out groups where either data_type has fewer than 2 observations
  filter(n_distinct(data_type) == 2) %>% 
  add_count(data_type) %>%
  filter(n >= 2) %>%
  # Run the test
  wilcox_test(
    Moran_residuals ~ data_type, 
    paired = FALSE
  ) %>%
  add_significance()


print(unpaired_results)
```

```{r}
row_idx_ns <- out_all$data_type == "nonspatial"
row_idx_s <- out_all$data_type == "spatial"
wilcox_df <- list()
for (fm in unique(out_all$fit_model)) {
  row_idx_f <- out_all$fit_model == fm
  for (dm in unique(out_all$data_model)) {
    row_idx_d <- out_all$data_model == dm
    
    s_ns <- out_all$Moran_residuals[row_idx_d & row_idx_f & row_idx_ns]
    s_s <- out_all$Moran_residuals[row_idx_d & row_idx_f & row_idx_s]
    
    if (0 == length(s_ns) || 0 == length(s_s)) {
      next
    }
    
    wt <- wilcox.test(s_ns, s_s, conf.int = TRUE)
    # Calculate Z from the W statistic
    # (This is an approximation used for large samples)
    n1 <- length(s_ns)
    n2 <-  length(s_s)
    mu_w <- (n1 * n2) / 2
    sigma_w <- sqrt((n1 * n2 * (n1 + n2 + 1)) / 12)
    z <- (wt$statistic - mu_w) / sigma_w
    
    r <- abs(z) / sqrt(n1 + n2)

    cd <- effsize::cliff.delta(s_s, s_ns)

    wilcox_df[[1 + length(wilcox_df)]] <- data.frame(
      data_model = dm, 
      fit_model = fm, 
      n_spatial = n2,
      n_nonspatial = n1, 
      w_stat = wt$statistic, 
      w_pval = wt$p.value,
      w_z = z,
      w_r = r,
      med_spatial = median(s_s),
      med_nonspatial = median(s_ns),
      hodges_lehmann = wt$estimate,
      cliffs_delta = cd$estimate,
      cliffs_delta_interp = cd$magnitude
    )
  }
}
wilcox_df <- dplyr::bind_rows(wilcox_df)
wilcox_df
```


```{r}
nonspat_summary_out$data_type <- "nonspatial"
summary_out$data_type <- "spatial"

plot_df <- merge(nonspat_summary_out[, cols_keep_out], 
                 summary_out[, cols_keep_out], by=cols_keep_out[1:5])

p <- ggplot2::ggplot(plot_df, ggplot2::aes(x=fit_model, y = Moran_residuals.y - Moran_residuals.x, fill = fit_model, color=fit_model))
p <- p + ggplot2::geom_boxplot(outliers=FALSE)
p <- p + ggplot2::scale_fill_manual(values = model_fill_vals)
p <- p + ggplot2::scale_color_manual(values = model_color_vals)
p <- p + ggplot2::scale_y_continuous(trans = signed_sqrt_trans)
p <- p + ggplot2::geom_hline(yintercept = 0, color = "red", linetype = "dashed", size = 1)
p <- p + ggplot2::labs(x = "Fitted Model", 
                       y = expression("Change in Moran's " * I * " of Residuals: " * I[Spatial] - I[Non-Spatial]), 
                       color = "Fitted Model", fill = "Fitted Model")
p <- p + ggplot2::theme_bw()
p <- p + ggplot2::theme(text = element_text(size=20), axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1), legend.position="none")
p
ggplot2::ggsave(paste0(figure_path, "morans_change_boxplot_", pattern_type, ".pdf"), p, width=12, height=9, units = "in", dpi = 300)
```

```{r}
# Difference in normalized MSE between spatial and non-spatial
nonspat_summary_out_all$data_model <- nonspat_summary_out_all$datagen_model
nonspat_summary_out_all$data_type <- "nonspatial"
summary_out_all$data_type <- "spatial"
nonspat_summary_out$data_type <- "nonspatial"
summary_out$data_type <- "spatial"
# Because trial is missing for the GLMs
summary_out_all[is.na(summary_out_all$trial), ]$trial <- summary_out_all[is.na(summary_out_all$trial), ]$trial_idx

plot_df <- merge(nonspat_summary_out_all[, cols_keep_out], 
                 summary_out_all[, cols_keep_out], by=cols_keep_out[1:5])

p1 <- ggplot2::ggplot(plot_df, ggplot2::aes(x = gene, 
                                            y = (MSE_counts_total.y / Mean_counts2_total.y) 
                                            - (MSE_counts_total.x / Mean_counts2_total.x), 
                                            fill = fit_model, color = fit_model))
p1 <- p1 + ggplot2::geom_boxplot(outliers = FALSE)
p1 <- p1 + ggplot2::geom_hline(yintercept = 0, color = "red", linetype = "dashed", size = 1)
p1 <- p1 + ggplot2::facet_wrap(~data_model)
p1 <- p1 + ggplot2::scale_color_manual(name = "", values = model_color_vals)
p1 <- p1 + ggplot2::scale_fill_manual(name = "", values = model_fill_vals)
p1 <- p1 + ggplot2::guides(fill = ggplot2::guide_legend(byrow = TRUE))
p1 <- p1 + ggplot2::guides(color = ggplot2::guide_legend(byrow = TRUE))
p1 <- p1 + ggplot2::facet_wrap(~paste0("Poisson-", data_model))
p1 <- p1 + ggplot2::scale_x_discrete(labels = c("Maximum", expression(Q[3]), "Median", expression(Q[1]), "Minimum"))
p1 <- p1 + ggplot2::scale_y_continuous(trans = signed_sqrt_trans)
p1 <- p1 + coord_cartesian(ylim = c(-1, 6))
p1 <- p1 + ggplot2::labs(x = "Gene Variance", y = expression("Change in Normalized MSE of Residuals: " * MSE[Spatial] - MSE[Non-Spatial]))
p1 <- p1 + ggplot2::theme_bw()
p1 <- p1 + ggplot2::theme(text = element_text(size=20),
                        legend.position = "bottom",
                        axis.text.y.right = element_blank(),
                        axis.line.y.right = element_blank(),
                        axis.ticks.y.right = element_blank())
ggplot2::ggsave(paste0(figure_path, "mse_change_boxplot.pdf"), p1, width=12, height=10, units = "in", dpi = 300)
```

```{r}
nonspat_summary_out$data_type <- "nonspatial"
summary_out$data_type <- "spatial"

plot_df <- merge(nonspat_summary_out[, cols_keep_out], 
                 summary_out[, cols_keep_out], by=cols_keep_out[1:5])

p <- ggplot2::ggplot(plot_df, ggplot2::aes(x = Moran_residuals.y - Moran_residuals.x, fill = fit_model))
p <- p + ggplot2::geom_density(alpha = 0.3)
p <- p + scale_fill_manual(values = model_fill_vals)
p <- p + ggplot2::scale_y_continuous(trans = signed_sqrt_trans)
p <- p + ggplot2::facet_wrap(~fit_model)
p <- p + ggplot2::labs(x = "Change in Moran's I of Residuals\n(Spatial - Non-Spatial)")
p <- p + ggplot2::theme_bw()
p <- p + ggplot2::theme(text = element_text(size=20), axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
p
ggplot2::ggsave(paste0(figure_path, "morans_change_density_", pattern_type, ".pdf"), p, width=12, height=9, units = "in", dpi = 300)
```

```{r, fig.width = 25, fig.height=20}
# Moran counts v. Moran residuals
p <- ggplot2::ggplot(plot_df, aes(x = Moran_counts.y - Moran_counts.x, y = Moran_residuals.y - Moran_residuals.x))
p <- p + ggplot2::geom_point(alpha = 0.5)
p <- p + ggplot2::facet_wrap(~fit_model)
p <- p + ggplot2::labs(x = "Change in Moran's I(counts)\n(Spatial - Non_spatial)", y = "Change in Moran's I(residuals)\n(Spatial - Non_spatial)")
p <- p + ggplot2::theme_bw()
p <- p + ggplot2::theme(text = element_text(size=20))
p
ggplot2::ggsave(paste0(figure_path, "morans_nonspatial_and_spatial_", pattern_type, ".pdf"), p, width=12, height=9, units = "in", dpi = 300)
```

```{r, fig.width = 25, fig.height=20}
# Moran counts v. Moran residuals
p <- ggplot2::ggplot(plot_df, ggplot2::aes(x = fit_model, y = (Moran_residuals.y - Moran_residuals.x)))
p <- p + ggplot2::geom_boxplot(outliers = FALSE)
p <- p + ggplot2::labs(x = "Model", y = "Change in Moran's I(residuals)\n(Spatial - Non_spatial)")
p <- p + ggplot2::theme_bw()
p <- p + ggplot2::theme(text = element_text(size=20))
p
```

```{r}
# Change in morans v. change in MSE
p <- ggplot2::ggplot(plot_df, ggplot2::aes(x =  MSE_counts_sample.y / Mean_counts2_sample.y 
                                           -  MSE_counts_sample.x / Mean_counts2_sample.x,
                                           y = Moran_residuals.y - Moran_residuals.x
                                           ))
p <- p + ggplot2::geom_point(alpha = 0.5)
p <- p + ggplot2::facet_wrap(~fit_model)
p <- p + ggplot2::xlim(-2, 2)
p <- p + ggplot2::labs(x = "Change in Relative MSE\n(Spatial - Non_spatial)", y = "Change in Moran's I(residuals)\n(Spatial - Non_spatial)")
p <- p + ggplot2::theme_bw()
p <- p + ggplot2::theme(text = element_text(size=20))
p
ggplot2::ggsave(paste0(figure_path, "mse_change_morans_spatial_", pattern_type, ".pdf"), p, width=12, height=9, units = "in", dpi = 300)
```

```{r}
p <- ggplot2::ggplot(plot_df, ggplot2::aes(x =  MSE_counts_sample.y / Mean_counts2_sample.y,
                                           y = Moran_residuals.y
                                           ))
p <- p + ggplot2::geom_point(alpha = 0.5)
p <- p + ggplot2::facet_wrap(~fit_model)
p <- p + ggplot2::xlim(-0.5, 2)
p <- p + ggplot2::labs(x = "Relative MSE (Spatial)", y = "Moran's I(residuals)(Spatial)")
p <- p + ggplot2::theme_bw()
p <- p + ggplot2::theme(text = element_text(size=20))
p
ggplot2::ggsave(paste0(figure_path, "mse_morans_spatial_", pattern_type, ".pdf"), p, width=12, height=9, units = "in", dpi = 300)
```


\newpage

# Error in gamma

Recall that $\gamma \in [0, 1)$ for the Leroux model.

Error in $\gamma$: $\hat{\gamma} - \gamma$

```{r}
if(pattern_type == "Leroux"){
  plot_data <- summary_out[which(summary_out$fit_model %in% c("TESSERA-Leroux")), ]
  plot_data <- merge(plot_data, pct_xpr_df, by = c("gene", "sample"), all.x=TRUE)
  
  p <- ggplot2::ggplot(plot_data, aes(x = sample, y = gamma_hat - gamma_true, color = gamma_true))
  p <- p + ggplot2::geom_boxplot(outliers = FALSE)
  p <- p + ggplot2::facet_wrap(~gene)
  p <- p + ggplot2::labs(x = "Sample", y = expression(paste("Error in ", gamma, ":   ", hat(gamma) - gamma)))
  p <- p + ggplot2::scale_color_viridis_c(name = expression(gamma), option = "turbo")
  p <- p + ggplot2::theme_bw()
  p <- p + ggplot2::theme(text = element_text(size=20))
  p <- p + ggplot2::theme(axis.text.x = element_text(size=12, angle = 90, hjust = 1, vjust = 0.5))
  p
  ggplot2::ggsave(paste0(figure_path, "synth_gamma_error.pdf"), p, width=8, height=6, units = "in", dpi = 300)
}
```

```{r}
if(pattern_type == "Leroux"){
  plot_data <- summary_out[which(summary_out$fit_model %in% c("TESSERA-Leroux")), ]
  plot_data <- merge(plot_data, pct_xpr_df, by = c("gene", "sample"), all.x=TRUE)
  
  p <- ggplot2::ggplot(plot_data, aes(x = sample, y = (gamma_hat - gamma_true) / gamma_true, color = gamma_true))
  p <- p + ggplot2::geom_boxplot(outliers = FALSE)
  p <- p + ggplot2::facet_wrap(~gene)
  p <- p + ggplot2::labs(x = "Sample", y = expression(paste("Relative error in ", gamma, ":   ", (hat(gamma) - gamma) / gamma)))
  p <- p + ggplot2::scale_color_viridis_c(name = expression(gamma), option = "turbo")
  p <- p + ggplot2::ylim(0, 1)
  p <- p + ggplot2::theme_bw()
  p <- p + ggplot2::theme(text = element_text(size=20))
  p <- p + ggplot2::theme(axis.text.x = element_text(size=12, angle = 90, hjust = 1, vjust = 0.5))
  p
  ggplot2::ggsave(paste0(figure_path, "synth_norm_gamma_error.pdf"), p, width=8, height=6, units = "in", dpi = 300)
}
```

\newpage

# Normalized error in tau^2

Recall that $\tau^2 \in [0, \infty)$ for all models.

Normalized error in $\tau^2$: $(\hat{\tau^2} - \tau^2) / \tau^2$

```{r}
if(pattern_type == "Leroux"){
  plot_data <- summary_out[which(summary_out$fit_model %in% c("TESSERA-Leroux")), ]
  plot_data <- merge(plot_data, pct_xpr_df, by = c("gene", "sample"), all.x=TRUE)
  p <- ggplot2::ggplot(plot_data, aes(x = sample, y = (tau2_hat - tau2_true), color = tau2_true))
  p <- p + ggplot2::geom_boxplot()
  p <- p + ggplot2::facet_wrap(~gene)
  p <- p + ggplot2::labs(x = "Sample", y = expression(paste("Error in ", tau^2, ":   ", widehat(tau^2) - tau^2)))
  p <- p + ggplot2::scale_color_viridis_c(name = expression(tau^2), option = "turbo")
  p <- p + ggplot2::theme_bw()
 p <- p + ggplot2::theme(text = element_text(size=20))
  p <- p + ggplot2::theme(axis.text.x = element_text(size=12, angle = 90, hjust = 1, vjust = 0.5))
  p
  ggplot2::ggsave(paste0(figure_path, "synth_tau2_error.pdf"), p, width=8, height=6, units = "in", dpi = 300)
}
```

```{r}
if(pattern_type == "Leroux"){
  plot_data <- summary_out[which(summary_out$fit_model %in% c("TESSERA-Leroux")), ]
  plot_data <- summary_out[which(!(summary_out$gene == "IGKC")), ]
  plot_data <- merge(plot_data, pct_xpr_df, by = c("gene", "sample"), all.x=TRUE)
  p <- ggplot2::ggplot(plot_data, aes(x = sample, y = (tau2_true - tau2_hat), color = tau2_true))
  p <- p + ggplot2::geom_boxplot()
  p <- p + ggplot2::facet_wrap(~gene)
  p <- p + ggplot2::labs(x = "Sample", y = expression(paste("Error in ", tau^2, ":   ", tau^2 - widehat(tau^2))))
  p <- p + ggplot2::scale_color_viridis_c(name = expression(tau^2), option = "turbo")
  p <- p + ggplot2::theme_bw()
 p <- p + ggplot2::theme(text = element_text(size=20))
  p <- p + ggplot2::theme(axis.text.x = element_text(size=12, angle = 90, hjust = 1, vjust = 0.5))
  p
  ggplot2::ggsave(paste0(figure_path, "synth_tau2_error_wo_IGKC.pdf"), p, width=8, height=6, units = "in", dpi = 300)
}
```

```{r}
if(pattern_type == "Leroux"){
  plot_data <- summary_out[which(summary_out$fit_model %in% c("TESSERA-Leroux")), ]
  plot_data <- merge(plot_data, pct_xpr_df, by = c("gene", "sample"), all.x=TRUE)
  p <- ggplot2::ggplot(plot_data, aes(x = sample, y = (tau2_hat - tau2_true) / tau2_true, color = tau2_true))
  p <- p + ggplot2::geom_boxplot(outliers = FALSE)
  p <- p + ggplot2::facet_wrap(~gene)
  p <- p + ggplot2::labs(x = "Sample", y = expression(paste("Relative error in ", tau^2, ": ", (widehat(tau)^2 - tau^2) / tau^2)))
  p <- p + ggplot2::scale_color_viridis_c(name = expression(tau^2), option = "turbo")
  p <- p + ggplot2::theme_bw()
 p <- p + ggplot2::theme(text = element_text(size=20))
  p <- p + ggplot2::theme(axis.text.x = element_text(size=12, angle = 90, hjust = 1, vjust = 0.5))
  p
  ggplot2::ggsave(paste0(figure_path, "synth_norm_tau2_error.pdf"), p, width=8, height=6, units = "in", dpi = 300)
}
```

```{r}
if(pattern_type == "Leroux"){
  plot_data <- summary_out[which(summary_out$gene == "IGKC" & summary_out$fit_model == "TESSERA-Leroux"), ]
  plot_data <- merge(plot_data, pct_xpr_df, by = c("gene", "sample"), all.x=TRUE)
  p <- ggplot2::ggplot(plot_data, aes(x = sample, y = (tau2_true - tau2_hat) / tau2_true, color = tau2_true))
  p <- p + ggplot2::geom_boxplot(outliers = FALSE)
  p <- p + ggplot2::facet_wrap(~gene)
  p <- p + ggplot2::labs(x = "Sample", y = expression(paste("Error in ", tau^2, ": ", (widehat(tau)^2 - tau^2) / tau^2)))
  p <- p + ggplot2::scale_color_viridis_c(name = expression(tau^2), option = "turbo")
  p <- p + ggplot2::theme_bw()
 p <- p + ggplot2::theme(text = element_text(size=20))
  p <- p + ggplot2::theme(axis.text.x = element_text(size=12, angle = 90, hjust = 1, vjust = 0.5))
  p
  ggplot2::ggsave(paste0(figure_path, "synth_norm_tau2_error_IGKC.pdf"), p, width=8, height=6, units = "in", dpi = 300)
}
```


\newpage

# Error in gamma and how it relates to tau^2

We conclude that the TESSERA model struggles to estimate low $\tau^2$ values. This also leads to worse estimates of $\gamma$.

```{r}
if(pattern_type == "Leroux"){
  plot_data <- summary_out[which(summary_out$fit_model %in% c("TESSERA-Leroux")), ]
  p <- ggplot2::ggplot(plot_data, aes(x = tau2_true, y = gamma_true - gamma_hat, color=sample))
  p <- p + ggplot2::geom_point(size=1)
  p <- p + scale_color_manual(values = sample_color_vals, name = "Sample")
  p <- p + ggplot2::geom_hline(yintercept = 0)
  p <- p + ggplot2::facet_wrap(~gene)
  p <- p + ggplot2::labs(x = expression(tau^2), y = expression(paste("Error in ", gamma, ":   ", gamma - hat(gamma))))
  p <- p + guides(color = guide_legend(override.aes = list(size = 5)))
  p <- p + ggplot2::theme_bw()
  p <- p + ggplot2::theme(text = element_text(size=20))
  p <- p + ggplot2::theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
  p
  ggplot2::ggsave(paste0(figure_path, "synth_tau2_error_gamma_error.pdf"), p, width=8, height=6, units = "in", dpi = 300)
}
```

\newpage

```{r}
if(pattern_type == "Leroux"){
  plot_data <- summary_out[which(summary_out$fit_model %in% c("TESSERA-Leroux")), ]
  p <- ggplot2::ggplot(plot_data, aes(x = gamma_true, y = (tau2_true - tau2_hat) / tau2_true, color=sample))
  p <- p + ggplot2::geom_point(size=1)
  p <- p + scale_color_manual(values = sample_color_vals, name = "Sample")
  p <- p + ggplot2::geom_hline(yintercept = 0)
  p <- p + ggplot2::facet_wrap(~gene)
  p <- p + ggplot2::labs(x = expression(gamma), y = expression(paste("Normalized error in ", tau^2, ":   ", (tau^2 - hat(tau^2)) / tau^2)))
  p <- p + guides(color = guide_legend(override.aes = list(size = 5)))
  p <- p + ggplot2::theme_bw()
  p <- p + ggplot2::theme(text = element_text(size=20))
  p <- p + ggplot2::theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
  p
  ggplot2::ggsave(paste0(figure_path, "synth_gamma_error_tau2_error.pdf"), p, width=8, height=6, units = "in", dpi = 300)
}
```

